---
title: "T-ALL WGS Nanopore Annotation Report"
format:
  dashboard:
    theme: cerulean
    orientation: rows
    scrolling: false
    embed-resources: true
knitr:
  opts_chunk:
    fig.align: 'center'
    tidy: TRUE
    fig.width: 7
    fig.height: 5
    R.options:
      stringsAsFactors: FALSE
params:
  meta_data: ""
  candidate_genes: ""
  ensembl_db: ""
  disco_enable: true
  proteinpaint_host: "http://localhost:3456"
  proteinpaint_genome: "hg38"
  # COHORT_MERGE_SV outputs. Empty on pass 1 of the two-pass render; the cohort
  # section then shows an explanatory card instead of the tables.
  cohort_recurrent_tsv: ""
  cohort_merged_vcf: ""
cache: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
.libPaths(c(getwd(),.libPaths()))

.pkg_lib <- file.path(getwd(), "R_runtime_libs")
dir.create(.pkg_lib, showWarnings = FALSE, recursive = TRUE)
.libPaths(c(.pkg_lib, .libPaths()))

.pkg_repos <- local({
  r <- getOption("repos")
  r <- r[!is.na(r) & nzchar(r) & r != "@CRAN@"]
  if (length(r) == 0) c(CRAN = "https://cloud.r-project.org")
  else unique(c(r, CRAN = "https://cloud.r-project.org"))
})

install_if_missing <- function(pkg, required = TRUE) {
  if (requireNamespace(pkg, quietly = TRUE)) return(invisible(TRUE))
  try(install.packages(pkg, repos = .pkg_repos, quiet = TRUE, lib = .pkg_lib),
      silent = TRUE)
  ok <- requireNamespace(pkg, quietly = TRUE)
  if (!ok && isTRUE(required)) {
    stop(sprintf(paste0(
      "Required R package '%s' is missing and could not be installed.\n",
      "  repos tried : %s\n",
      "  lib target  : %s\n",
      "This host has no working route to a package repository. Either bake '%s'\n",
      "into the SV_REPORT_INDEX container image (preferred - see the TODO below),\n",
      "or export http_proxy/https_proxy for the task so the runtime install works."),
      pkg, paste(.pkg_repos, collapse = ", "), .pkg_lib, pkg), call. = FALSE)
  }
  invisible(ok)
}
install_if_missing("DT")
install_if_missing("plotly")

# Runtime Bioconductor install — used for the optional EnsDb breakend
# annotation block. Rocker/verse ships a P3M repos config so BiocManager
# prefers binaries when available, keeping cold install closer to seconds
# than compile-from-source minutes. Wrapped in tryCatch so a repo miss or
# transient network hiccup falls back to the downstream `requireNamespace`
# gate rather than killing the whole render.
# TODO: bake ensembldb + plyranges into a custom SV_REPORT_INDEX container
# image so this runtime install can be removed.
install_bioc_if_missing <- function(pkg) {
  if (requireNamespace(pkg, quietly = TRUE)) return(invisible(TRUE))
  tryCatch({
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager", repos = .pkg_repos,
                       quiet = TRUE, lib = .pkg_lib)
    }
    BiocManager::install(pkg, update = FALSE, ask = FALSE, quiet = TRUE,
                         lib = .pkg_lib)
    invisible(TRUE)
  }, error = function(e) {
    message("Optional Bioc install skipped for '", pkg, "': ",
            conditionMessage(e))
    invisible(FALSE)
  })
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(DT)
  library(plotly)
})

#  Reusable predicates / normalizers (tidyverse-patterns) 
# True iff x has real annotation content (handles NA, "", ".", "NA", "()").
is_meaningful <- function(x) {
  !is.na(x) & !(x %in% c("", ".", "NA", "()"))
}

# Strip InfoTuple-style "(...)" / quote decoration from a vembrane TSV cell and
# return the first element as a clean string. Splits on EITHER `,` or `;`. The `,` handles raw vembrane InfoTuple input like "('5', 'full=5')" while `;` handles the output like "5; full=5; full=5"
# scalar_first applied value returns the entire joined string and downstream as.integer() / factor() match returns NA.
scalar_first <- function(x) {
  cleaned <- str_replace_all(as.character(x), "[()'\"\\s]", "")
  str_split_fixed(cleaned, "[,;]", 2)[, 1]
}

# Cast ACMG_class to an ordered factor for plot ordering and color mapping.
# NA-like inputs (R-NA, empty string, ".", literal "NA") are normalized to the
# explicit "NA" factor level so plots can render them as grey via the
# acmg_palette mapping below (true-NA factor values would render transparent
# under scale_fill_viridis_d).
acmg_factor <- function(x) {
  s <- scalar_first(x)
  s[is.na(s) | s %in% c("", ".", "NA")] <- "NA"
  factor(s, levels = c("0", "1", "2", "3", "4", "5", "NA"))
}

# Strict ACMG cleaner for the cohort TSV: returns integer 1-5 or NA_integer_.
# Coerces 0 (AnnotSV "unclassified") and any non-1-5 value to NA. Designed to
# overwrite the raw ACMG_class column in the cohort TSV — NOT a replacement
# for acmg_factor(), which intentionally keeps "0" / "NA" as factor levels
# for plot color mapping.
acmg_clean <- function(x) {
  s <- scalar_first(x)
  n <- suppressWarnings(as.integer(s))
  ifelse(is.na(n) | n < 1L | n > 5L, NA_integer_, n)
}

# Color palette: viridis "rocket" ramp for ACMG classes 0–5, grey for "NA".
# Used by Plot 2 (ACMG class stacked bars) via scale_fill_manual so the
# NA category renders as grey instead of transparent.
acmg_palette <- c(
  setNames(viridisLite::rocket(6, direction = -1, end = 0.85),
           c("0", "1", "2", "3", "4", "5")),
  "NA" = "grey70"
)

#  BND partner-chromosome validation for ST17 hits 
# AnnotSV's user-BED overlap test only checks one breakpoint of a BND
# record against the BED interval.
#
# Current ST17 BED `name` column format (post gene-annotation update):
#   "chrA:posA-chrB:posB_(geneA::geneB)"                (single hit)
#   "hit1|hit2|hit3|..."                                (multi-hit, '|'-sep)
#

# Strip a leading "chr" (any case) so "chr21", "Chr21", "21" all compare equal.
normalize_chrom <- function(s) {
  sub("^chr", "", as.character(s), ignore.case = TRUE)
}

# Strip InfoTuple wrapping `('val',)` -> `val`. Pass-through for already-clean strings.
strip_infotuple <- function(s) {
  s <- as.character(s)
  s <- sub("^\\(['\"]", "", s)
  s <- sub("['\"],?\\s*\\)$", "", s)
  s
}

# Clean vembrane's TSV-encoded ClinVar_SV_CLNSIG cell into a semicolon-joined
# list of unique significance labels. VEP's --custom for the ClinVar SV VCF emits per-overlap CLNSIG values wrapped in double quotes and joined with "&", e.g. `"Pathogenic"&"Pathogenic"&"Likely%20pathogenic"&"Benign"`. 
clean_clinvar_clnsig <- function(x) {
  if (is.na(x) || !nzchar(x)) return("")
  s <- str_replace_all(as.character(x), '"', '')
  parts <- str_split(s, "&")[[1]] |> str_trim()
  parts <- parts[nzchar(parts)]
  paste(unique(parts), collapse = "; ")
}

# Vectorized breakpoint formatter for the DT `breakpoint` column.
#
# Intra-chromosomal SVs / CNAs (DEL, DUP, INV, INS, CNV):
#   "chr21:100000-110000"   (if END > POS)
#   "chr21:100000"          (if END == POS or END missing)
#
# Interchromosomal BNDs (parses partner from ALT bracket notation, all
# four VCF spec forms: t[p[ , [p[t , t]p] , ]p]t):
#   "chr21:100000|chr5:300000"
#
# BND with unparseable ALT falls back to "chr21:100000".
#
# Fully vectorized via `str_match` on the ALT column (no rowwise). The
# `chr` prefix is added when absent (raw CHROM in the merged VCF is bare
# "21" while ALT partner uses "chr21" — normalize both sides for consistent
# display).
compute_breakpoint <- function(chrom, pos, end, alt, svtype) {
  chrom  <- as.character(chrom)
  pos    <- as.character(pos)
  end    <- as.character(end)
  alt    <- as.character(alt)
  svtype <- as.character(svtype)

  prefix_chr <- function(c) ifelse(is.na(c) | !nzchar(c) | startsWith(c, "chr"),
                                   c, paste0("chr", c))

  chrom <- prefix_chr(chrom)

  # BND if SVTYPE says so OR if ALT has bracket notation
  is_bnd <- (svtype == "BND") | grepl("[\\[\\]]", alt)

  # Vectorized partner parse from bracket notation
  m <- str_match(alt, "[\\[\\]]([^:\\[\\]]+):([0-9]+)[\\[\\]]")
  p_chr <- prefix_chr(m[, 2])
  p_pos <- m[, 3]
  partner_ok <- !is.na(p_chr) & !is.na(p_pos)

  # Intra-chromosomal display: "chr:pos-end" or "chr:pos" when end is
  # missing / equals pos (typical for INS)
  end_valid <- !is.na(end) & nzchar(end) & end != "." & end != pos
  intra <- ifelse(end_valid,
                  paste0(chrom, ":", pos, "-", end),
                  paste0(chrom, ":", pos))

  ifelse(
    is_bnd & partner_ok,
    paste0(chrom, ":", pos, "|", p_chr, ":", p_pos),
    intra
  )
}

# Round a comma-joined hVAF cell to 2 decimal places.
# Input:  "0.800000011920929,0.0,0.0"  (HP1, HP2, unphased triple)
# Output: "0.80,0.00,0.00"
# Preserves the 3-value tuple shape; missing/empty inputs pass through.
format_hvaf <- function(x, digits = 2L) {
  if (is.na(x) || !nzchar(x)) return("")
  parts <- str_split(as.character(x), ",")[[1]] |> str_trim()
  parts <- vapply(parts, function(v) {
    n <- suppressWarnings(as.numeric(v))
    if (is.na(n)) v else formatC(n, format = "f", digits = digits)
  }, character(1), USE.NAMES = FALSE)
  paste(parts, collapse = ",")
}

# check if a bnd is intra or interchromosomal 
chk_intra_bnd <- function(breakpoint){
  be <- str_split_1(breakpoint, pattern = "\\|") |>  gsub(pattern = "(.+):[0-9]+", replacement = "\\1", x = _)
  if(length(be) == 2){
    intrachrom <- be[1] == be[2]
  } else {
    intrachrom <- TRUE
  }
  return(intrachrom)
}

# Summarize an AnnotSV P_*_source (P_gain_source / P_loss_source /
# P_ins_source) cell into a compact "first-3 + count" per category form.
#
# Input format: the raw INFO field is Number=. String — one entry per
# overlapping annotated interval, each entry pipe-separated hits with the
# source prefix. vembrane's chr(44).join(InfoTuple) delivers the whole cell
# as ONE string with commas between entries and pipes within entries, e.g.:
#   "dbVar:nssv1|dbVar:nssv2|morbid:APP|morbid:BACH1,.,.,dbVar:nssv2|morbid:APP,."
#
# Output form (per cell):
#   "dbVar[3/12]: nssv1|nssv2|nssv3; morbid[3/8]: APP|BACH1|JAM2;
#    CLN[2]: 3779864|552391; HI3[1]: DYRK1A"
# For dbVar and morbid we show first-3 hits + total (unique) count. For
# CLN and HI3 we show first-3 hits + count too. Bare `nssv*` (no prefix)
# is categorized as dbVar per AnnotSV convention. Empty / all-"." input
# returns "".
summarize_p_source <- function(x, top_n = 3L) {
  if (is.na(x) || !nzchar(x)) return("")
  # Explode across commas (Number=. entries) and pipes (per-hit)
  parts <- str_split(as.character(x), "[,|]")[[1]] |> str_trim()
  parts <- parts[nzchar(parts) & parts != "."]
  if (length(parts) == 0) return("")
  # Categorize
  cat_of <- function(h) {
    if (startsWith(h, "dbVar:") || startsWith(h, "nssv")) return("dbVar")
    if (startsWith(h, "morbid:")) return("morbid")
    if (startsWith(h, "CLN:"))    return("CLN")
    if (startsWith(h, "HI3:"))    return("HI3")
    if (startsWith(h, "TS3:"))    return("TS3")
    "other"
  }
  # Strip prefix for display
  strip_prefix <- function(h) sub("^[A-Za-z0-9]+:", "", h)

  cats  <- vapply(parts, cat_of, character(1), USE.NAMES = FALSE)
  bags  <- split(parts, cats)
  order_shown <- c("dbVar", "morbid", "CLN", "HI3", "TS3", "other")
  order_shown <- order_shown[order_shown %in% names(bags)]

  fmt_one <- function(cat) {
    uniq <- unique(bags[[cat]])
    shown <- head(vapply(uniq, strip_prefix, character(1), USE.NAMES = FALSE), top_n)
    sprintf("%s[%d/%d]: %s", cat, length(shown), length(uniq),
            paste(shown, collapse = "|"))
  }
  paste(vapply(order_shown, fmt_one, character(1), USE.NAMES = FALSE),
        collapse = "; ")
}

# Clean AnnotSV's `full,split,split,...` padding pattern. When an AnnotSV
# TSV row is converted to VCF INFO, the full-mode value is at index 0 and
# the split-mode entries follow at positions 1..N. Fields that vary per
# split (Gene_name, Location, etc.) get real values; fields that only exist
# in full mode get "." for every split → strings like "value,.,.,.,.".
# Split on ",", drop "." and empty tokens, rejoin with ",". Idempotent —
# a cell that never had padding (already clean) round-trips unchanged.
clean_annot_padding <- function(x) {
  if (is.na(x) || !nzchar(x)) return("")
  parts <- str_split(as.character(x), ",")[[1]] |> str_trim()
  parts <- parts[nzchar(parts) & parts != "."]
  paste(parts, collapse = ",")
}

# Split on both `,` and `|`, drop empties / ".", return unique tokens in
# first-seen order joined with a single delimiter. Used for AnnotSV fields
# where full-mode emits a `|`-joined multi-value list and split-mode repeats
# the same values comma-appended (Gene_name, CytoBand, Location).
dedup_annot_multi <- function(x, sep = "; ") {
  if (is.na(x) || !nzchar(x)) return("")
  parts <- str_split(as.character(x), "[,|]\\s*")[[1]] |> str_trim()
  parts <- parts[nzchar(parts) & parts != "."]
  if (length(parts) == 0) return("")
  paste(unique(parts), collapse = sep)
}

# Collapse a multi-cytoband CytoBand string to its encompassing range.
# For large CNAs / SVs, AnnotSV emits both a "full" annotation with the
# encompassing range (e.g. `q23.31-q26.3`) and per-split-mode annotations
# for every sub-band inside (`q23.31`, `q23.32`, `q23.32-q23.33`, …). After
# `dedup_annot_multi()` runs, the CytoBand cell looks like:
#     q23.31-q26.3; q23.31; q23.32; q23.32-q23.33; q23.33; q24.1; ...
# which is technically deduplicated (every distinct sub-range is unique)
# but useless in the DT display. Collapse to a single `<min>-<max>` range
# by parsing every position token (splitting ranges at `-`), sorting by
# (arm, major, sub) ascending, and rebuilding the outermost interval.
#
# Input examples → outputs:
#   "q23.31"                                    → "q23.31"
#   "q23.31; q23.32"                            → "q23.31-q23.32"
#   "q23.31-q26.3; q23.31; ...; q26.3"          → "q23.31-q26.3"
#   "p13.2; p13.1"                              → "p13.1-p13.2"   (p arm order)
#   ""  / "." / NA                              → ""
#
# p-arm bands are sorted with the LARGER major number distal (further from
# centromere) — i.e. p13 comes before p11 when reading toward the centromere.
# For range endpoints we still take alphanumeric min-max of the numeric part
# on each arm; the display convention is "smaller number - larger number"
# per arm, matching how AnnotSV writes its own full-mode ranges.
collapse_cytoband_range <- function(cytoband_str) {
  if (is.na(cytoband_str) || !nzchar(cytoband_str) || cytoband_str == ".") return("")
  parts <- str_split(as.character(cytoband_str), "[;,|]\\s*")[[1]] |> str_trim()
  parts <- parts[nzchar(parts) & parts != "."]
  if (length(parts) == 0) return("")

  # Break any pre-existing ranges (e.g. "q23.31-q26.3") into their two
  # endpoint tokens so we sort on a flat vector of atomic positions.
  positions <- unlist(lapply(parts, function(p) {
    if (grepl("-", p, fixed = TRUE)) str_split(p, "-")[[1]] else p
  }))
  positions <- unique(positions[nzchar(positions)])
  if (length(positions) == 0) return("")
  if (length(positions) == 1) return(positions[1])

  # Sort into chromosomal (linear) order — p-telomere → centromere → q-telomere.
  # On the p arm, HIGHER band numbers are DISTAL (closer to telomere, i.e. lower
  # linear position): p13.3 < p13.2 < p13.1 in chromosomal position; AnnotSV
  # writes p-arm ranges telomere-to-centromere → descending numeric.
  # On the q arm, HIGHER band numbers are DISTAL (further from centromere, i.e.
  # higher linear position): q11.1 < q11.2 < q13.3 in chromosomal position;
  # AnnotSV writes q-arm ranges centromere-to-telomere → ascending numeric.
  # Sub-band digit strings compare LEXICOGRAPHICALLY (not numerically) since
  # they're nested labels: q26.1 contains q26.11 / q26.12 / q26.13, and q26.2
  # comes AFTER q26.13. Lex "13" < "3" because '1' < '3' — correct order.
  parse_key <- function(p) {
    m <- regmatches(p, regexec("^([pq])(\\d+)(?:\\.(\\d+))?$", p))[[1]]
    if (length(m) < 3 || !nzchar(m[2])) return(list(arm=3L, major=999L, sub="~"))
    arm_rank <- if (m[2] == "p") 1L else 2L
    major <- suppressWarnings(as.integer(m[3])) %||% 999L
    sub   <- if (length(m) >= 4 && nzchar(m[4])) m[4] else ""
    list(arm = arm_rank, major = major, sub = sub)
  }
  keys <- lapply(positions, parse_key)
  arms   <- vapply(keys, `[[`, integer(1),   "arm")
  majors <- vapply(keys, `[[`, integer(1),   "major")
  subs   <- vapply(keys, `[[`, character(1), "sub")

  # Sort each arm independently, then concat p-then-q to get chromosomal order.
  #   p arm: sort ascending (major, sub) then REVERSE — high band numbers first
  #          (most telomeric on p side)
  #   q arm: sort ascending (major, sub) directly
  is_p <- arms == 1L
  is_q <- arms == 2L
  is_x <- arms == 3L  # unrecognized tokens sink to the end
  p_ord <- rev(order(majors[is_p], subs[is_p]))
  q_ord <- order(majors[is_q], subs[is_q])
  x_ord <- order(subs[is_x])
  sorted <- c(positions[is_p][p_ord],
              positions[is_q][q_ord],
              positions[is_x][x_ord])
  min_p <- sorted[1]
  max_p <- sorted[length(sorted)]
  if (min_p == max_p) return(min_p)
  paste0(min_p, "-", max_p)
}

# Parse a CIViC INFO annotation cell into a fixed set of interpretive columns.
# The CIViC VCF annotation loaded via `bcftools annotate` uses a VEP-style
# 40-field pipe-delimited entry; a record with N CIViC assertions carries N
# such entries joined by ",". This helper extracts the seven columns
# requested for downstream clinical review, deduplicates across per-record
# entries (multi-therapy / multi-evidence records collapse cleanly), and
# returns a named list of atomic character scalars suitable for
# `dplyr::mutate(...)` splatting. Field indices are fixed by the CIViC
# header (see the CIVIC INFO Description in the annotate_civic_snv VCF):
#   0 Allele        1 Consequence   2 SYMBOL          9 CIViC Variant ID
#  11 CIViC Variant URL             24 CIViC Entity Variant Origin
#  29 CIViC Entity Therapies        34 CIViC Assertion ACMG Codes
parse_civic_field <- function(civic_str) {
  empty <- list(
    CIViC_Consequence           = NA_character_,
    CIViC_SYMBOL                = NA_character_,
    CIViC_Variant_ID            = NA_character_,
    CIViC_Variant_URL           = NA_character_,
    CIViC_Entity_Variant_Origin = NA_character_,
    CIViC_Entity_Therapies      = NA_character_,
    CIViC_Assertion_ACMG_Codes  = NA_character_
  )
  s <- as.character(civic_str)
  if (is.na(s) || !nzchar(s) || s %in% c("NA", ".")) return(empty)

  hits <- str_split(s, ",")[[1]]
  fields_list <- lapply(hits, function(h) {
    v <- str_split(h, "\\|")[[1]]
    length(v) <- 40L
    v
  })

  # 0-indexed in the CIViC header; +1 for R.
  extract_col <- function(idx0) {
    vals <- vapply(fields_list, function(v) {
      x <- v[idx0 + 1L]
      if (is.na(x) || !nzchar(x)) NA_character_ else x
    }, character(1))
    vals <- unique(vals[!is.na(vals)])
    if (length(vals) == 0) NA_character_ else paste(vals, collapse = ";")
  }

  list(
    CIViC_Consequence           = extract_col(1L),
    CIViC_SYMBOL                = extract_col(2L),
    CIViC_Variant_ID            = extract_col(9L),
    CIViC_Variant_URL           = extract_col(11L),
    CIViC_Entity_Variant_Origin = extract_col(24L),
    CIViC_Entity_Therapies      = extract_col(29L),
    CIViC_Assertion_ACMG_Codes  = extract_col(34L)
  )
}


# Vectorized full clean of vembrane InfoTuple representations across a column.
clean_tuple <- function(x) {
  s <- as.character(x)
  out <- s
  hit <- !is.na(s) & grepl("^\\(.*\\)$", s)
  if (any(hit)) {
    out[hit] <- vapply(s[hit], function(t) {
      # Strip outer parens (and trailing comma for single-element tuples)
      inner <- sub("^\\(", "", sub(",?\\s*\\)$", "", t))
      # Capture all single- or double-quoted elements
      m <- str_match_all(inner, "'([^']*)'|\"([^\"]*)\"")[[1]]
      if (nrow(m) == 0) return(inner)
      vals <- ifelse(is.na(m[, 2]), m[, 3], m[, 2])
      paste(vals, collapse = "; ")
    }, character(1), USE.NAMES = FALSE)
  }
  out
}

# Load a one-gene-per-line file into a character vector for Candidate_genes
# Returns an empty vector if the params$candidate_genes path is missing or empty.
load_candidate_gene_set <- function() {
  fp <- params$candidate_genes
  if (is.null(fp) || !nzchar(fp) || !file.exists(fp)) return(character(0))
  lines <- readLines(fp, warn = FALSE)
  trimmed <- str_trim(lines)
  trimmed[nzchar(trimmed) & !startsWith(trimmed, "#")]
}

# Union candidate-gene detection across Gene_name (SV interval overlap) +
# Closest_left / Closest_right (nearest flanking gene). Intergenic BND
# breakpoints have empty Gene_name but often carry a driver gene as the
# closest-flanking annotation — critical for T-ALL where BND-mediated fusion
# partners land in intergenic regions. Returns a "; "-joined string of unique
# candidate-gene hits across the three columns, in first-seen order.
compute_candidate_genes <- function(gene_name, closest_left, closest_right,
                                    candidate_set) {
  if (length(candidate_set) == 0) return("")
  raw <- c(gene_name, closest_left, closest_right)
  raw <- raw[!is.na(raw) & nzchar(raw)]
  if (length(raw) == 0) return("")
  parts <- unlist(lapply(raw, function(s) str_trim(str_split(s, "[|;]\\s*")[[1]])))
  hits  <- unique(parts[parts %in% candidate_set])
  paste(hits, collapse = "; ")
}

# Summarize a Gene_name list for display: if the underlying gene count
# exceeds `cap`, truncate to the first `cap` genes and append " (n=<count>
# genes)"; otherwise return the raw joined string unchanged. `count` should
# be the AnnotSV Gene_count column value (authoritative — includes genes
# that may have been de-duplicated in the display string). Falls back to
# counting the delimiter-split length when count is NA / empty.
summarize_gene_list <- function(gene_str, count = NA_integer_, cap = 20L) {
  if (is.na(gene_str) || !nzchar(gene_str)) return("")
  genes <- str_split(gene_str, "[|;]\\s*")[[1]] |> str_trim()
  genes <- genes[nzchar(genes)]
  n_parsed <- length(genes)
  n_auth <- suppressWarnings(as.integer(count))
  n_show <- if (!is.na(n_auth) && n_auth > 0) n_auth else n_parsed
  if (n_show > cap) {
    paste0(paste(head(genes, cap), collapse = "; "),
           "; ... (n=", n_show, " genes)")
  } else {
    paste(genes, collapse = "; ")
  }
}

# Parse an ST17 score cell (semicolon- or comma-joined multi-hit) and return
# the maximum integer score. Falls back to 0 for empty / non-numeric input.
# Used as the tier-1 tie-breaker in driver_tier ranking (recurrence-count
# proxy: highest ST17 score = most recurrent T-ALL SV).
parse_st17_max_score <- function(score_str) {
  if (is.na(score_str) || !nzchar(score_str)) return(0L)
  toks <- str_split(score_str, "[,;|]\\s*")[[1]] |> str_trim()
  vals <- suppressWarnings(as.integer(toks))
  vals <- vals[!is.na(vals)]
  if (length(vals) == 0) 0L else max(vals)
}

# Pair-split an ST17 (name, score) cell together so downstream ordering is
# stable and interpretable. AnnotSV emits multi-hit ST17 cells with two
# different join-levels: the name field can hold multiple gene-pair
# alternatives inside a single BED row's `_(GENE_A|_GENE_B::GENE_C|_GENE_D)`
# construct AND multiple BED rows joined by `|`; the score field carries one
# score per BED row joined by `|`. That mismatch is why a length-5 name and
# length-2 score can appear in the same record. This function:
#   1. Splits name via `split_pipe_outside_parens` (respects `_(...|...)`).
#   2. Splits score by `|` at outer level.
#   3. Drops empty/"." tokens (residual `clean_annot_padding` artefacts).
#   4. Pairs by index — if scores shorter than names (AnnotSV pipe-mismatch),
#      the LAST score is repeated to fill (a per-BED-row score naturally
#      applies to every gene-pair variant inside that BED row's name field).
#   5. De-duplicates by name (keeps highest-score occurrence).
#   6. Orders by score desc (NA last) and truncates to `top_n` hits.
# Returns a named list of two `|`-joined strings so downstream code can
# either display them jointly (`ST17_hit`) or feed the score string into
# `parse_st17_max_score` for tier tie-breaking.
parse_st17_pair <- function(name_str, score_str, top_n = 3L) {
  empty_result <- list(names = "", scores = "")
  if (is.na(name_str) || !nzchar(name_str) || name_str == ".") return(empty_result)

  names_v <- split_pipe_outside_parens(as.character(name_str))
  names_v <- names_v[nzchar(names_v) & names_v != "."]
  if (length(names_v) == 0) return(empty_result)

  scores_num <- if (is.na(score_str) || !nzchar(score_str) || score_str == ".") {
    integer(0)
  } else {
    toks <- str_split(as.character(score_str), "\\|")[[1]] |> str_trim()
    toks <- toks[nzchar(toks) & toks != "."]
    suppressWarnings(as.integer(toks))
  }
  scores_num <- scores_num[!is.na(scores_num)]

  n_names  <- length(names_v)
  n_scores <- length(scores_num)
  if (n_scores == 0) {
    paired <- rep(NA_integer_, n_names)
  } else if (n_scores >= n_names) {
    paired <- scores_num[seq_len(n_names)]
  } else {
    paired <- c(scores_num, rep(scores_num[n_scores], n_names - n_scores))
  }

  ord <- order(paired, decreasing = TRUE, na.last = TRUE)
  names_v <- names_v[ord]
  paired  <- paired[ord]

  keep <- !duplicated(names_v)
  names_v <- names_v[keep]
  paired  <- paired[keep]

  n_take <- min(as.integer(top_n), length(names_v))
  list(
    names  = paste(names_v[seq_len(n_take)], collapse = "|"),
    scores = paste(ifelse(is.na(paired[seq_len(n_take)]), "",
                          as.character(paired[seq_len(n_take)])),
                   collapse = "|")
  )
}

# Assign a T-ALL driver-priority tier per pass2 record:
#   Tier 1 = any SVTYPE-matched name_ST17_* hit (empirical T-ALL recurrence)
#   Tier 2 = no ST17 hit, but ≥1 Candidate_genes match (literature prior)
#   Tier 3 = neither (still passed pass2 by other evidence — P_gain_source,
#                     RE_gene, ACMG 3-5, or ClinVar pathogenic)
# Tie-breaker `tier_score` (higher = better):
#   T1: max integer score across SVTYPE-matched score_ST17_* — proxies
#       T-ALL cohort recurrence
#   T2: number of overlapping candidate genes — more hits = broader impact
#   T3: AnnotSV_ranking_score — falls back to AnnotSV's own ranking
# Returns a tibble with columns `driver_tier` (integer 1-3) and
# `tier_score` (numeric).
assign_driver_tier <- function(st17_name, st17_score, candidate_genes,
                                annotsv_ranking_score) {
  has_st17 <- !is.na(st17_name) & nzchar(st17_name)
  has_cand <- !is.na(candidate_genes) & nzchar(candidate_genes)
  tier <- ifelse(has_st17, 1L,
                 ifelse(has_cand, 2L, 3L))
  cand_n <- vapply(candidate_genes, function(s) {
    if (is.na(s) || !nzchar(s)) return(0L)
    length(str_split(s, ";\\s*")[[1]])
  }, integer(1), USE.NAMES = FALSE)
  st17_max <- vapply(st17_score, parse_st17_max_score,
                     integer(1), USE.NAMES = FALSE)
  ann_num <- suppressWarnings(as.numeric(annotsv_ranking_score))
  ann_num[is.na(ann_num)] <- 0
  score <- ifelse(tier == 1L, as.numeric(st17_max),
                  ifelse(tier == 2L, as.numeric(cand_n), ann_num))
  tibble::tibble(driver_tier = tier, tier_score = score)
}

# CNA driver-priority tier 
# BED CNV 
#   Tier 1 = any SVTYPE-matched name_ST19_* hit 
#            match GAIN records -> AMP+GAIN cols, LOSS records -> DEL+LOSS)
#   Tier 2 = no ST19 hit, but ≥1 Candidate_genes match
#   Tier 3 = neither (P_gain_source, ACMG 3-5, etc.)
# Tie-breaker `tier_score`:
#   T1: max score across the 4 score_ST19_* columns
#   T2: number of overlapping candidate genes
#   T3: AnnotSV_ranking_score
assign_driver_tier_cna <- function(st19_amp_name, st19_del_name,
                                    st19_gain_name, st19_loss_name,
                                    st19_amp_score, st19_del_score,
                                    st19_gain_score, st19_loss_score,
                                    candidate_genes, annotsv_ranking_score) {
  st19_names <- list(st19_amp_name, st19_del_name, st19_gain_name, st19_loss_name)
  st19_scores <- list(st19_amp_score, st19_del_score, st19_gain_score, st19_loss_score)

  has_st19 <- Reduce(`|`, lapply(st19_names, function(v) !is.na(v) & nzchar(v)))
  has_cand <- !is.na(candidate_genes) & nzchar(candidate_genes)
  tier <- ifelse(has_st19, 1L,
                 ifelse(has_cand, 2L, 3L))

  cand_n <- vapply(candidate_genes, function(s) {
    if (is.na(s) || !nzchar(s)) return(0L)
    length(str_split(s, ";\\s*")[[1]])
  }, integer(1), USE.NAMES = FALSE)

  st19_max <- pmax(
    vapply(st19_amp_score,  parse_st17_max_score, integer(1), USE.NAMES = FALSE),
    vapply(st19_del_score,  parse_st17_max_score, integer(1), USE.NAMES = FALSE),
    vapply(st19_gain_score, parse_st17_max_score, integer(1), USE.NAMES = FALSE),
    vapply(st19_loss_score, parse_st17_max_score, integer(1), USE.NAMES = FALSE),
    na.rm = TRUE
  )
  ann_num <- suppressWarnings(as.numeric(annotsv_ranking_score))
  ann_num[is.na(ann_num)] <- 0
  score <- ifelse(tier == 1L, as.numeric(st19_max),
                  ifelse(tier == 2L, as.numeric(cand_n), ann_num))
  tibble::tibble(driver_tier = tier, tier_score = score)
}

# SNV driver-priority tier 
# candidate-gene evidence.
#   Tier 1 = ONC = "Oncogenic" OR "Likely_oncogenic" AND SYMBOL in candidate panel
#   Tier 2 = ClinVar pathogenic/likely_pathogenic OR candidate-gene member
#   Tier 3 = neither ( SIFT deleterious / PolyPhen damaging predictions) 
# Tie-breaker `tier_score` = max(CLAIRSTO_VAF, DEEPSOMATIC_VAF) — somatic. ties broken by higher confidence in the somatic call.
assign_driver_tier_snv <- function(onc, clnsig, candidate_genes,
                                    clairsto_vaf, deepsomatic_vaf) {
  onc_str    <- as.character(onc)
  clnsig_str <- as.character(clnsig)
  cand_str   <- as.character(candidate_genes)

  # ONC substring: "Oncogenic" and "Likely_oncogenic"
  # ClinVar substring: "Pathogenic", "Likely_pathogenic", "Pathogenic_low_penetrance"
  has_onc  <- !is.na(onc_str) &
              grepl("ncogenic", onc_str, ignore.case = TRUE)
  has_cln  <- !is.na(clnsig_str) &
              grepl("athogenic", clnsig_str, ignore.case = TRUE)
  has_cand <- !is.na(cand_str) & nzchar(cand_str)

  tier <- ifelse(has_onc & has_cand, 1L,
                 ifelse(has_cln | has_cand, 2L, 3L))

  vaf <- pmax(
    suppressWarnings(as.numeric(clairsto_vaf)),
    suppressWarnings(as.numeric(deepsomatic_vaf)),
    na.rm = TRUE
  )
  vaf[is.na(vaf) | is.infinite(vaf)] <- 0
  tibble::tibble(driver_tier = tier, tier_score = vaf)
}

# Split on '|' but only at paren depth 0 — the ST17 hit format embeds gene
# annotations like `_(RUNX1|_TLX3::TLX3|_RUNX1)` where the inner '|' is
# part of one hit, not a hit-separator. `str_split` doesn't natively
# respect bracket nesting; this manual scanner does.
split_pipe_outside_parens <- function(s) {
  if (length(s) == 0 || is.na(s) || s == "") return(character(0))
  chars <- strsplit(s, "", fixed = TRUE)[[1]]
  depth <- 0L
  parts <- character(0)
  current <- ""
  for (ch in chars) {
    if (ch == "(") { depth <- depth + 1L; current <- paste0(current, ch) }
    else if (ch == ")") { depth <- max(depth - 1L, 0L); current <- paste0(current, ch) }
    else if (ch == "|" && depth == 0L) {
      parts <- c(parts, current); current <- ""
    } else {
      current <- paste0(current, ch)
    }
  }
  c(parts, current)
}

# Parse the partner-side breakpoint from a Severus/SAVANA BND ALT.
# Handles all four VCF spec bracket forms: t[p[ , [p[t , t]p] , ]p]t.
parse_bnd_partner <- function(alt) {
  if (is.na(alt) || alt == "") return(list(chrom = NA_character_, pos = NA_real_))
  m <- str_match(alt, "[\\[\\]]([^:\\[\\]]+):([0-9]+)[\\[\\]]")
  if (is.na(m[1, 2])) return(list(chrom = NA_character_, pos = NA_real_))
  list(chrom = normalize_chrom(m[1, 2]), pos = as.numeric(m[1, 3]))
}

# Test whether a sample's two breakpoints overlap a reference pair within
# `tol` bp, trying both pair orderings (BND pairs are unordered).
bnd_pair_matches <- function(s_chrom_a, s_pos_a, s_chrom_b, s_pos_b,
                              r_chrom_a, r_pos_a, r_chrom_b, r_pos_b,
                              tol = 10000) {
  s_a <- normalize_chrom(s_chrom_a); s_b <- normalize_chrom(s_chrom_b)
  r_a <- normalize_chrom(r_chrom_a); r_b <- normalize_chrom(r_chrom_b)
  direct  <- (s_a == r_a && abs(s_pos_a - r_pos_a) <= tol &&
              s_b == r_b && abs(s_pos_b - r_pos_b) <= tol)
  swapped <- (s_a == r_b && abs(s_pos_a - r_pos_b) <= tol &&
              s_b == r_a && abs(s_pos_b - r_pos_a) <= tol)
  isTRUE(direct || swapped)
}

# Apply the partner filter to one sample BND row. Returns the filtered
# `name_ST17_BND` string (sub-elements separated by '|' that don't
# partner-match the sample dropped). `score_ST17_BND` is passed through
# unchanged: AnnotSV's score join uses '|' at a different level than its
# name-join (one BED row's `name` field can itself contain '|' between
# gene-pair variants), so a 1:1 name↔score index alignment isn't safe.
# If the partner cannot be parsed from ALT, clear the name (conservative
# — a loose-end breakend can't be validated).
filter_bnd_st17_name <- function(sample_chrom, sample_pos, sample_alt,
                                  ref_names, tol = 10000) {
  cleaned <- strip_infotuple(ref_names)
  if (is.na(cleaned) || cleaned %in% c("", "NA", ".")) return("")

  partner <- parse_bnd_partner(sample_alt)
  if (is.na(partner$chrom)) return("")

  hits <- split_pipe_outside_parens(cleaned)

  # Per-hit format: chrA:posA-chrB:posB_(genes...)
  # Match the two coord pairs separated by '-'; anything after '_' is
  # gene-annotation suffix and ignored for partner validation.
  keep <- map_lgl(hits, function(hit) {
    m <- str_match(hit, "([^:|]+):([0-9]+)-([^:|]+):([0-9]+)")
    if (is.na(m[1, 2])) return(FALSE)
    bnd_pair_matches(
      sample_chrom, as.numeric(sample_pos), partner$chrom, partner$pos,
      m[1, 2], as.numeric(m[1, 3]),
      m[1, 4], as.numeric(m[1, 5]),
      tol = tol
    )
  })

  paste(hits[keep], collapse = "|")
}
```

```{r annotsv-user-cols-regex}
#  Derive the user-annotation column regex from AnnotSV header files 
# ANNOTSV_SETUPUSERANNO stages `<STn>_<basename>.header.tsv` files under
# annotsv_annotations/Annotations_Human/Users/<genome>/{FtIncludedInSV,
# SVincludedInFt,AnyOverlap}/. vembrane emits the corresponding INFO fields
# as `name_<STn>_<basename>` and `score_<STn>_<basename>` in the pipeline
# TSVs, so the "Custom Annot. Hits" tally in parse_annotsv() must only
# count columns backed by actual staged headers, not any incidental `_ST`
# prefix. `build_user_cols_regex()` reads those filenames at render time
# and builds the exact alternation regex; when the annotsv_annotations
# input isn't staged (e.g. dev-time render outside the pipeline), fall
# back to the structural `^(name|score)_ST\d+_` prefix pattern which still
# beats the pre-refactor `FtIncludedInSV|SVincludedInFt|AnyOverlap` regex.
build_user_cols_regex <- function() {
  users_root <- file.path("annotsv_annotations", "Annotations_Human", "Users")
  if (!dir.exists(users_root)) return("^(name|score)_ST\\d+_")
  hdr_files <- list.files(
    path       = users_root,
    pattern    = "\\.header\\.tsv$",
    recursive  = TRUE,
    full.names = FALSE
  )
  if (length(hdr_files) == 0) return("^(name|score)_ST\\d+_")
  basenames <- unique(sub("\\.header\\.tsv$", "", basename(hdr_files)))
  # Regex-escape literal chars (dot, paren, plus, etc.) so `.` in a
  # basename like `Alterations.CNV.Recurrent_AMP` matches a literal dot.
  escaped <- gsub("([][{}()*+?.\\\\^$|])", "\\\\\\1", basenames)
  sprintf("^(name|score)_(%s)$", paste(escaped, collapse = "|"))
}

user_cols_regex <- build_user_cols_regex()
```

```{r dt-readability, results='asis'}
# Cyborg theme renders the page in dark mode (black body, light grey text).
# DT tables inherit that styling and become unreadable (black on dark grey).
# Force the table cell area to white with dark text while keeping the
# surrounding chrome (filter / info / pagination / column-search) light so
# it stays readable against the dark page background.
cat('
<style>
  /*  Layout: width is handled per-DT-chunk via `#| column: screen-inset`
     (Quarto chunk option). This widens those specific chunks beyond the
     centered article body without overriding Bootstrap container widths
     elsewhere (which is fragile). Inside each chunk, force the DT widget
     and the table itself to fill 100% of the screen-inset region.  */
  .column-screen-inset .dataTables_wrapper,
  .column-screen-inset .datatables,
  .column-screen-inset .html-widget,
  .column-screen-inset table.dataTable { width: 100% !important; }

  /*  Typography: smaller section headers  */
  h1, .h1 { font-size: 1.6rem !important; }
  h2, .h2 { font-size: 1.3rem !important; margin-top: 1.2rem; }
  h3, .h3 { font-size: 1.1rem !important; margin-top: 1.0rem; }
  h4, .h4 { font-size: 1.0rem !important; }

  /*  Typography: smaller text inside DT tables  */
  table.dataTable, table.dataTable thead th, table.dataTable tbody td,
  .dataTables_wrapper, .dataTables_filter, .dataTables_length,
  .dataTables_info, .dataTables_paginate, .dt-buttons .dt-button {
    font-size: 0.80rem !important;
  }
  table.dataTable thead th, table.dataTable tbody td { padding: 4px 6px !important; }

  /*  Cell wrapping: allow long gene lists / coordinate strings to wrap
     instead of bulldozing horizontal layout. max-width caps cell width
     while word-break lets multi-element values break at any character.  */
  table.dataTable tbody td {
    white-space: normal !important;
    word-break: break-word !important;
    overflow-wrap: anywhere !important;
    max-width: 320px;
    vertical-align: top !important;
  }
  table.dataTable thead th {
    white-space: normal !important;
    vertical-align: bottom !important;
  }

  /*  Column grid lines: a 1 px right border on every td/th keeps headers
     visually aligned over their column values when the table wraps long
     strings into multi-line cells. Last column drops the right border so
     the table edge stays clean. Header row gets a heavier bottom border
     so the head/body boundary remains clear under the cyborg dark theme.  */
  table.dataTable td, table.dataTable th {
    border-right: 1px solid #ced4da !important;
  }
  table.dataTable td:last-child, table.dataTable th:last-child {
    border-right: none !important;
  }
  table.dataTable thead th {
    border-bottom: 2px solid #495057 !important;
  }

  /*  Theme: cell backgrounds (light) on the dark cyborg body  */
  table.dataTable { background-color: #ffffff !important; color: #212529 !important; }
  table.dataTable thead th,
  table.dataTable thead td { background-color: #e9ecef !important; color: #212529 !important;
                             border-bottom: 2px solid #adb5bd !important; }
  table.dataTable tbody tr td { background-color: #ffffff !important; color: #212529 !important; }
  table.dataTable tbody tr.odd td,
  table.dataTable.stripe tbody tr.odd > td,
  table.dataTable.display tbody tr.odd > td { background-color: #f1f3f5 !important; }
  table.dataTable tbody tr:hover td { background-color: #dee2e6 !important; }
  /* Per-column search row inside thead */
  table.dataTable thead input,
  table.dataTable thead select { background-color: #ffffff !important; color: #212529 !important;
                                 border: 1px solid #ced4da !important; }

  /* Wrapper chrome — keep light text against the dark page */
  .dataTables_wrapper { color: #f8f9fa !important; padding: 4px 0; }
  .dataTables_filter label,
  .dataTables_length label,
  .dataTables_info { color: #f8f9fa !important; }
  .dataTables_filter input,
  .dataTables_length select { background-color: #ffffff !important; color: #212529 !important;
                              border: 1px solid #ced4da !important; }
  .dataTables_paginate .paginate_button { color: #f8f9fa !important; }
  .dataTables_paginate .paginate_button.current,
  .dataTables_paginate .paginate_button:hover { background: #495057 !important;
                                                color: #ffffff !important; border-color: #6c757d !important; }
  /* CSV / Excel export buttons */
  .dt-buttons .dt-button { background-color: #495057 !important; color: #ffffff !important;
                          border: 1px solid #6c757d !important; }
</style>
')
```

```{r parse-annotsv}
annotsv_tsvs <- list.files(pattern = "\\.merged\\.annotated.*\\.tsv$",
                           full.names = TRUE)
# Exclude pass1/pass2 TSVs (also match the .tsv$ pattern but belong to
# the vembrane-table chunk below).
annotsv_tsvs <- setdiff(annotsv_tsvs,
                        list.files(pattern = "\\.filtered\\.pass[12]\\.tsv$",
                                   full.names = TRUE))

parse_annotsv <- function(tsv_path) {
  df   <- read_tsv(tsv_path, col_types = cols(.default = "c"), show_col_types = FALSE)
  full <- filter(df, Annotation_mode == "full")
  sample_id <- sub("\\.merged\\..+$", "", basename(tsv_path))
  acmg <- suppressWarnings(as.integer(full$ACMG_class))
  # `user_cols_regex` is built from the staged AnnotSV Users/ header files
  # (see annotsv-user-cols-regex chunk) so only actual user-annotation
  # columns count toward `Custom Annot. Hits`.
  user_cols <- grep(user_cols_regex, colnames(full), value = TRUE)
  custom_hits <- if (length(user_cols) > 0) {
    rowSums(!is.na(full[user_cols]) & full[user_cols] != "", na.rm = TRUE) > 0
  } else {
    rep(FALSE, nrow(full))
  }
  tibble(
    `Sample ID`             = sample_id,
    `Total SVs`             = nrow(full),
    `ACMG >= VUS (>=3)`     = sum(acmg >= 3, na.rm = TRUE),
    `Likely Pathogenic (4)` = sum(acmg == 4, na.rm = TRUE),
    `Pathogenic (5)`        = sum(acmg == 5, na.rm = TRUE),
    `Custom Annot. Hits`    = sum(custom_hits)
  )
}

stats_df <- map_dfr(annotsv_tsvs, parse_annotsv)
```

```{r merge-metadata}
# Join key is `id` (the pipeline meta.id, e.g. "colo829"), not `sample_id`
# (which is a friendly label like "sample_1" in the metadata CSV). `Sample ID`
# in stats_df is derived from the AnnotSV TSV filename and therefore matches
# `id` exactly. Strip leading/trailing whitespace from metadata column names —
# the test fixture has " mutations" (with a leading space) due to a CSV
# whitespace artifact.
metadata_file <- params$meta_data
if (!is.null(metadata_file) && nzchar(metadata_file) && file.exists(metadata_file)) {
  metadata <- read_csv(metadata_file, col_types = cols(.default = "c"),
                       show_col_types = FALSE)
  colnames(metadata) <- str_trim(colnames(metadata))
  stats_df <- left_join(stats_df, metadata, by = c("Sample ID" = "id"))
} else {
  metadata <- NULL
}

# Helper: attach sample_metadata rows to a variant table (SV / CNA / SNV) by
# joining on the pipeline id (`SAMPLE` in the variant df, `id` in the
# metadata CSV). Called twice per variant table:
#   * `cols = NULL` — full metadata merge, used before `write_tsv` so the
#     cohort TSVs carry every metadata column.
#   * `cols = "sample_id"` — DT-only merge to add the friendly sample_id
#     label without cluttering the interactive table with all cohort fields.
# Returns df unchanged when no metadata is available so downstream code is
# always safe to call.
merge_sample_metadata <- function(df, meta = metadata, sample_col = "SAMPLE",
                                   cols = NULL) {
  if (is.null(meta) || nrow(meta) == 0 || !sample_col %in% names(df)) return(df)
  m <- if (is.null(cols)) meta
       else dplyr::select(meta, id, dplyr::any_of(cols))
  m <- m |> dplyr::distinct(id, .keep_all = TRUE)
  df |> dplyr::left_join(m, by = setNames("id", sample_col))
}
```

```{r link-columns}
knot_files <- list.files("knot", pattern = "\\.html$", full.names = FALSE)
xl_files   <- list.files("knot", pattern = "\\.xlsm$", full.names = FALSE)
vep_files  <- list.files("vep",  pattern = "\\.html$", full.names = FALSE)
igv_files  <- list.files("igv",  pattern = "\\.html$", full.names = FALSE)

# Bucketed IGV reports: IGV_REPORTS_SV emits one HTML per (sample × bucket)
# where bucket is {small, medium, large} is encoded in the filename.
# full `<sample>.pass2.igv.<bucket>` prefix so each bucket gets its own link.
make_link <- function(sample_id, files, subdir, label = "open &#x2197;",
                       suffix = NULL) {
  pat <- if (is.null(suffix)) {
    paste0("^", sample_id, "\\.")
  } else {
    paste0("^", sample_id, "\\.", suffix)
  }
  fname <- grep(pat, files, value = TRUE)
  if (length(fname) >= 1)
    sprintf('<a href="%s/%s" target="_blank">%s</a>', subdir, fname[1], label)
  else
    "-"
}

stats_df <- stats_df |>
  mutate(
    `KnotAnnotSV Report`   = map_chr(`Sample ID`, make_link, knot_files, "../annotations_sv/knotannotsv/html"),
    `KnotAnnotSV XLSM`     = if (length(xl_files) > 0)
                                map_chr(`Sample ID`, make_link, xl_files,
                                        "../annotations_sv/knotannotsv/xl",
                                        label = "download &#x2B07;")
                              else
                                "-",
    `VEP (breakpoint)`     = if (length(vep_files) > 0)
                                map_chr(`Sample ID`, make_link, vep_files, "../annotations_sv/ensemblvep/vep")
                              else
                                "-",
    `IGV small (<1 kb)`    = if (length(igv_files) > 0)
                                map_chr(`Sample ID`, make_link, igv_files, "../igv_reports",
                                        suffix = "pass2\\.igv\\.small")
                              else
                                "-",
    `IGV medium (1-10 kb)` = if (length(igv_files) > 0)
                                map_chr(`Sample ID`, make_link, igv_files, "../igv_reports",
                                        suffix = "pass2\\.igv\\.medium")
                              else
                                "-",
    `IGV large (>=10 kb)`  = if (length(igv_files) > 0)
                                map_chr(`Sample ID`, make_link, igv_files, "../igv_reports",
                                        suffix = "pass2\\.igv\\.large")
                              else
                                "-"
  )
```


```{r data-dict-loader, message = FALSE, warning = FALSE}
# Load per-variant-type VCF header dictionaries emitted by CREATE_DATA_DICT
# (one TSV per sample per variant type, from FILTER_{SV,CNA,SNV}). Union
# across samples and dedup by (field_type, field_name) — descriptions are
# stable across samples because they come from the same VCF-header block
# for a given caller + AnnotSV run.
load_data_dict <- function(subdir) {
  if (!dir.exists(subdir)) return(tibble::tibble())
  files <- Sys.glob(file.path(subdir, "*.data_dict.tsv"))
  if (length(files) == 0) return(tibble::tibble())
  dplyr::bind_rows(lapply(files, readr::read_tsv,
                          show_col_types = FALSE,
                          col_types = "cccccc")) |>
    dplyr::distinct(field_type, field_name, .keep_all = TRUE)
}

dict_sv  <- load_data_dict("data_dict_sv")
dict_cna <- load_data_dict("data_dict_cna")
dict_snv <- load_data_dict("data_dict_snv")

# Hand-curated overlay for columns computed inside this qmd (not present
# in the VCF header). Overlay takes precedence over any VCF-header
# definition of the same name — the qmd-side meaning is the authoritative
# one for reviewers reading the DT tables. Do NOT paraphrase VCF-header
# entries here; if the truthful description lives in the header, let the
# dict TSV supply it.
qmd_computed_defs <- tibble::tribble(
  ~field_name,        ~description,
  "SAMPLE",           "VCF sample-column value (from the source caller output; may differ from cohort sample_id if the caller renamed the sample)",
  "sample_id",        "Cohort-facing sample identifier joined in from the params.sample_metadata CSV; NA when no metadata table was supplied",
  "driver_tier",      "Prioritization tier: T1 = empirical T-ALL recurrence (ST17/ST19) OR oncogenic in the candidate panel; T2 = candidate-gene hit; T3 = other pass2 evidence",
  "tier_score",       "Within-tier tie-breaker (recurrence count, candidate-gene count, or AnnotSV_ranking_score depending on tier)",
  "breakpoint",       "Compact chr:pos display of the record's primary breakpoint (SV/BND) or interval start (CNA)",
  "CytoBand_display", "chr<CHROM> prefix prepended to the AnnotSV CytoBand cell for a full-locus label",
  "Gene_name_display","Semicolon-joined unique SYMBOL list, dedup'd from AnnotSV's full-mode + per-split-mode Gene_name concatenation",
  "Candidate_genes",  "Intersection of Gene_name with the user candidate-gene panel (params.candidate_genes)",
  "has_mate",         "Whether both ends of a BND event survived the filter (TRUE = mate pair present, FALSE = orphaned BND)",
  "ST17_hit",         "One-line summary of the highest-scored SVTYPE-matched ST17 T-ALL recurrence BED hit(s)",
  "ST17_name",        "Top-3 ST17 hit names (by score), pipe-outside-parens joined",
  "ST17_score",       "Top-3 ST17 hit scores, paired positionally with ST17_name",
  "ST19_hit",         "One-line summary of the top 3 SVTYPE-matched ST19 T-ALL recurrent-CNV BED hits (AMP / DEL / GAIN / LOSS union, ordered by score desc; format: \"hit1 [s1]; hit2 [s2]; hit3 [s3]\")",
  "ST19_name",        "Top-3 ST19 hit names (score-ordered), pipe-joined; consolidated from name_ST19_Alterations.CNV.Recurrent_{AMP,DEL,GAIN,LOSS}",
  "ST19_score",       "Top-3 ST19 hit scores, paired positionally with ST19_name; consolidated from score_ST19_Alterations.CNV.Recurrent_{AMP,DEL,GAIN,LOSS}",
  "CNA_TYPE",         "Wakhan copy-number call type derived from CN1/CN2 (GAIN, LOSS, CNLOH); drives tier assignment and colour palette",
  "WAKHAN_TYPE",      "Raw Wakhan classification (integer/integer/cnloh/…), preserved from the caller",
  "CIViC_Consequence","VEP Consequence value from the CIViC INFO tag — compare to the top-level Consequence",
  "CIViC_SYMBOL",     "Gene symbol from the CIViC INFO tag — compare to the top-level SYMBOL",
  "CIViC_Variant_ID", "CIViC internal variant identifier",
  "CIViC_Variant_URL","Deep link to the variant page on civicdb.org",
  "CIViC_Entity_Variant_Origin", "CIViC assertion of variant origin (somatic / germline / unknown)",
  "CIViC_Entity_Therapies",      "Therapies linked to the variant per CIViC evidence entries",
  "CIViC_Assertion_ACMG_Codes",  "ACMG/AMP variant-classification codes attached to the CIViC assertion"
)

# Emit a small download-button HTML linking to the full data-dictionary TSV
# for a given variant type. Relative path resolves against index.html once
# both files are published under results/sv_report/ by publishDir.
# Returns "" when no dict TSV is present so the caller can `cat()` it
# unconditionally without leaking placeholders into the report.
dict_download_button <- function(subdir, label) {
  if (!dir.exists(subdir)) return("")
  files <- Sys.glob(file.path(subdir, "*.data_dict.tsv"))
  if (length(files) == 0) return("")
  # Use the first (should be the only) file. Relative link so it works both
  # in the raw index.html and via any static-file server.
  rel <- files[1]
  sprintf('<p style="margin:8px 0 4px 0;">
  <a href="%s" download
     style="display:inline-block; padding:8px 14px;
            background:#0d6efd; color:#ffffff; text-decoration:none;
            border-radius:5px; font-size:0.9em; font-weight:500;">
    &#x2B07;&#xFE0F; Download full %s data dictionary (TSV)
  </a>
  <span style="font-size:0.82em; color:#666; margin-left:8px;">
    Every VCF INFO / FORMAT / FILTER declaration for this variant type.
    Join by field_name to the cohort_%s.tsv output for column definitions.
  </span>
</p>
', rel, label, tolower(label))
}

# Render a compact "column definition" table for a display data frame.
# Priority: qmd-computed overlay first; then VCF-header dict lookup;
# blank when neither has a description (surfaces the gap to reviewers
# so it can be filled in the overlay).
render_data_dict <- function(display_df, dict_tbl, caption = NULL) {
  cols <- names(display_df)
  desc <- vapply(cols, function(c) {
    hit_ov <- qmd_computed_defs$description[qmd_computed_defs$field_name == c]
    if (length(hit_ov) > 0 && nzchar(hit_ov[1])) return(hit_ov[1])
    if (nrow(dict_tbl) > 0) {
      hit <- dict_tbl$description[dict_tbl$field_name == c]
      if (length(hit) > 0 && nzchar(hit[1])) return(hit[1])
    }
    ""
  }, character(1))
  ftype <- vapply(cols, function(c) {
    if (nrow(dict_tbl) > 0) {
      hit <- dict_tbl$field_type[dict_tbl$field_name == c]
      if (length(hit) > 0) return(hit[1])
    }
    ""
  }, character(1))
  ftype[!nzchar(ftype) & nzchar(desc)] <- "computed"
  ftype[!nzchar(ftype)]                <- "unknown"

  DT::datatable(
    tibble::tibble(Column = cols, Kind = ftype, Definition = desc),
    caption   = caption,
    filter    = "none", rownames = FALSE, width = "100%",
    options   = list(dom = "t", pageLength = 100, ordering = FALSE,
                     autoWidth = TRUE,
                     columnDefs = list(list(width = "18%", targets = 0),
                                       list(width = "10%", targets = 1)))
  )
}
```

```{r valuebox-helpers, include = FALSE}
# Value-box helper. Every KPI card uses vb_count(df_name, cond_fn) which
# guards against the referenced data frame being missing (branch turned off
# in the pipeline) or empty — yields 0 instead of throwing. cond_fn is a
# function of the data frame returning a logical vector; sum() collapses.
vb_count <- function(df_name, cond_fn) {
  if (!exists(df_name)) return(0L)
  df <- get(df_name)
  if (!is.data.frame(df) || nrow(df) == 0) return(0L)
  tryCatch(as.integer(sum(cond_fn(df), na.rm = TRUE)),
           error = function(e) 0L)
}

# Defaults so the pure-markdown DISCO block on the Overview page can resolve
# its inline `r ...` references even when disco-plot-prep / disco-plot-data
# didn't fire (small cohorts with 0 Tier-1 records, or `disco_enable = FALSE`).
# These are OVERRIDDEN by the DISCO data chunks below when they do run —
# `<-` in a later chunk simply rebinds the name in the shared session.
pp_host           <- if (!is.null(params$proteinpaint_host)   && nzchar(params$proteinpaint_host))   params$proteinpaint_host   else "http://localhost:3456"
pp_genome         <- if (!is.null(params$proteinpaint_genome) && nzchar(params$proteinpaint_genome)) params$proteinpaint_genome else "hg38"
disco_mlst_json   <- "[]"
safe_cohort_label <- "Cohort"
n_cohort_snv      <- 0L
n_cohort_cna      <- 0L
n_cohort_bnd      <- 0L
n_cohort_samples  <- 0L
sample_ids        <- character(0)
lollipop_link_display <- tibble::tibble()
```





```{r load-pass2}
# Load candidate-gene set unconditionally so downstream CNA / SNV chunks
# always see `candidate_set` — even when the SV pass2-clean chunk (which
# also assigns it) is skipped because pass2_long is empty.
candidate_set <- load_candidate_gene_set()

# SVTYPE / CNA_TYPE color palette (colorblind-safe Dark2 hues). Defined
# here unconditionally so the CNA pass2 DT chunk can reuse it even when
# the SV pass2-table chunk is skipped due to zero SV records.
svtype_colors <- c(
  BND = "#7570B3",  # purple
  DEL = "#D95F02",  # orange
  DUP = "#1B9E77",  # teal
  INS = "#E7298A",  # magenta
  INV = "#66A61E"   # green
)

pass2_files <- list.files(pattern = "\\.filtered\\.pass2\\.tsv$", full.names = TRUE)

# Also accept the synthetic test fixture
pass2_files <- c(pass2_files,
                 list.files(pattern = "vembrane_pass2_syndata\\.tsv$", full.names = TRUE))

pass2_long <- if (length(pass2_files) > 0) {
  map_dfr(pass2_files, ~ read_tsv(.x, col_types = cols(.default = "c"),
                                  show_col_types = FALSE))
} else {
  tibble()
}

# vembrane's long-format prepends a SAMPLE column with the sample name from VCF
# FORMAT. Fall back to a filename-derived id if SAMPLE is absent (older outputs).
if (nrow(pass2_long) > 0 && !"SAMPLE" %in% names(pass2_long)) {
  pass2_long <- pass2_long |> mutate(SAMPLE = "unknown")
}

# Coerce numerics needed by plots
if (nrow(pass2_long) > 0) {
  pass2_long <- pass2_long |>
    mutate(
      ACMG_class_clean = acmg_factor(ACMG_class),
      ranking_score    = suppressWarnings(as.numeric(scalar_first(AnnotSV_ranking_score)))
    )
}
```

```{r pass2-clean}
#| eval: !expr nrow(pass2_long) > 0

# Moved up from the DT chunk so plot-st17 counts and the sample-stats
# `Custom Annot. Hits` tally see the same partner-validated ST17 columns
# as the DT. The order matters:
#   1. Strip vembrane InfoTuple wrapping
#   2. Compute Candidate_genes from Gene_name 
#   3. Cast ACMG_class to integer 1-5
#   4. convert ST17 name/score cells whose SVTYPE suffix doesn't match the
#      record's SVTYPE 
#   5. BND partner-chromosome filter: keep ST17 BND name sub-elements
#      where both reference breakpoints match (within 10 kb).
pass2_long <- pass2_long |>
  mutate(across(where(is.character), clean_tuple))

# Collapse AnnotSV's `full,split,split,...` padding on every character
# column (see clean_annot_padding docs). Runs after clean_tuple so any
# InfoTuple decoration is stripped first.
pass2_long <- pass2_long |>
  mutate(across(where(is.character), \(v) vapply(v, clean_annot_padding, character(1), USE.NAMES = FALSE)))

candidate_set <- load_candidate_gene_set()
# Union candidate-gene detection across Gene_name + Closest_left + Closest_right.
# The two Closest_* columns capture nearest-flanking genes when the SV falls
# in intergenic space — essential for BND breakpoints whose fusion partner
# is a driver gene that the SV interval itself doesn't overlap.
if (length(candidate_set) > 0 && "Gene_name" %in% names(pass2_long)) {
  closest_l <- if ("Closest_left"  %in% names(pass2_long)) pass2_long$Closest_left  else rep("", nrow(pass2_long))
  closest_r <- if ("Closest_right" %in% names(pass2_long)) pass2_long$Closest_right else rep("", nrow(pass2_long))
  pass2_long <- pass2_long |>
    mutate(Candidate_genes = pmap_chr(
      list(Gene_name, closest_l, closest_r),
      function(gn, cl, cr) compute_candidate_genes(gn, cl, cr, candidate_set)
    ))
} else {
  pass2_long$Candidate_genes <- ""
}

pass2_clean <- pass2_long |>
  mutate(ACMG_class = acmg_clean(ACMG_class)) |>
  select(-any_of("Annotation_mode"))

# Uniquify the ClinVar_SV_CLNSIG cell so a row overlapping 50 ClinVar records
# with the same "Pathogenic" label shows a single "Pathogenic" instead of a
# 50-element repeat wall.
if ("ClinVar_SV_CLNSIG" %in% names(pass2_clean)) {
  pass2_clean <- pass2_clean |>
    mutate(ClinVar_SV_CLNSIG = vapply(ClinVar_SV_CLNSIG, clean_clinvar_clnsig,
                                     character(1), USE.NAMES = FALSE))
}

# Round hVAF (HP1, HP2, unphased triple) to 2 decimal places so the DT reads
# "0.80,0.00,0.00" instead of "0.800000011920929,0.0,0.0".
if ("hVAF" %in% names(pass2_clean)) {
  pass2_clean <- pass2_clean |>
    mutate(hVAF = vapply(hVAF, format_hvaf, character(1), USE.NAMES = FALSE))
}

# breakpoint column: "chr21:100000-110000" for intra-chromosomal, or
# "chr21:100000|chr5:300000" for interchromosomal BNDs.
if (all(c("CHROM", "POS", "ALT", "SVTYPE") %in% names(pass2_clean))) {
  end_col <- if ("END" %in% names(pass2_clean)) pass2_clean$END else pass2_clean$POS
  pass2_clean <- pass2_clean |>
    mutate(breakpoint = compute_breakpoint(CHROM, POS, end_col, ALT, SVTYPE))
}

# Blank ST17 cells (name_ST17_* and score_ST17_*) whose suffix mismatches the record's SVTYPE. cur_column() extracts the suffix 
# convert the BND hits that intrachromosomal to be the intformation from the ST17 DEL, since ST17 BND is only interchromosomal translocations per authors definitions
pass2_clean <- pass2_clean |>
  mutate(intrachomosomal = purrr::map_lgl(.x = breakpoint, .f = ~chk_intra_bnd(.x))) |>
  mutate(name_ST17_BND = case_when(
    SVTYPE == "BND" & intrachomosomal == TRUE ~ name_ST17_DEL,
    TRUE ~ name_ST17_BND
  ), 
  score_ST17_BND = case_when(
    SVTYPE == "BND" & intrachomosomal == TRUE ~ score_ST17_DEL,
    TRUE ~ score_ST17_BND
  )) %>% 
  mutate(across(
    matches("^(name|score)_ST17_(BND|DEL|DUP|INS|INV)$"),
    ~ ifelse(SVTYPE == sub("^(name|score)_ST17_", "", cur_column()), as.character(.x), "")
  ))

# BND partner-chromosome filter (see helper docs in setup chunk).
if (all(c("name_ST17_BND", "CHROM", "POS", "ALT", "SVTYPE") %in% names(pass2_clean))) {
  pass2_clean <- pass2_clean |>
    mutate(name_ST17_BND = pmap_chr(
      list(SVTYPE, CHROM, POS, ALT, name_ST17_BND),
      function(svt, chrom, pos, alt, nm) {
        if (svt != "BND") return(as.character(nm))
        filter_bnd_st17_name(chrom, pos, alt, nm)
      }
    ))
}

#  Driver-tier ranking + display-only columns 
# Compute one collapsed ST17 hit + score per row (SVTYPE-matched — the
# other 4 name_ST17_/score_ST17_ columns are already blanked by the earlier
# suffix-mismatch mutate). ST17_hit is empty when no ST17 evidence exists.
st17_name_col <- function(svt) {
  switch(svt,
         BND = "name_ST17_BND", DEL = "name_ST17_DEL",
         DUP = "name_ST17_DUP", INS = "name_ST17_INS",
         INV = "name_ST17_INV", NA_character_)
}
st17_score_col <- function(svt) sub("^name_", "score_", st17_name_col(svt))

pick_st17 <- function(df, picker) {
  vapply(seq_len(nrow(df)), function(i) {
    col <- picker(df$SVTYPE[i])
    if (is.na(col) || !(col %in% names(df))) return("")
    val <- df[[col]][i]
    if (is.na(val)) "" else as.character(val)
  }, character(1))
}

pass2_clean <- pass2_clean |>
  mutate(
    ST17_name_raw  = pick_st17(pick(everything()), st17_name_col),
    ST17_score_raw = pick_st17(pick(everything()), st17_score_col)
  )

# Pair-split name + score together, dedup, order by score desc, truncate to
# top 3 hits. Both output columns keep the same ordering so the k-th name
# always pairs with the k-th score.
st17_pairs <- purrr::map2(pass2_clean$ST17_name_raw, pass2_clean$ST17_score_raw,
                          ~ parse_st17_pair(.x, .y, top_n = 3L))

pass2_clean <- pass2_clean |>
  mutate(
    ST17_name  = vapply(st17_pairs, `[[`, character(1), "names"),
    ST17_score = vapply(st17_pairs, `[[`, character(1), "scores"),
    # ST17_hit renders the top-3 pairs as "hit1 [s1]; hit2 [s2]; hit3 [s3]".
    # Blank when no ST17 evidence exists. When a score is missing (padded
    # NA slot from AnnotSV pipe-mismatch), the score bracket is omitted for
    # that entry.
    ST17_hit = vapply(seq_along(ST17_name), function(i) {
      nm <- ST17_name[i]; sc <- ST17_score[i]
      if (!nzchar(nm)) return("")
      nms <- split_pipe_outside_parens(nm)
      scs <- if (nzchar(sc)) str_split(sc, "\\|")[[1]] else rep("", length(nms))
      length(scs) <- length(nms)
      scs[is.na(scs)] <- ""
      pieces <- vapply(seq_along(nms), function(j) {
        if (nzchar(scs[j])) paste0(nms[j], " [", scs[j], "]") else nms[j]
      }, character(1))
      paste(pieces, collapse = "; ")
    }, character(1))
  ) |>
  select(-ST17_name_raw, -ST17_score_raw)

tier_tbl   <- assign_driver_tier(
  pass2_clean$ST17_name,
  pass2_clean$ST17_score,
  pass2_clean$Candidate_genes,
  pass2_clean$AnnotSV_ranking_score
)
pass2_clean <- bind_cols(pass2_clean, tier_tbl)

# Tier-3 MINDA-only filter. T3 records passed the pass2 gate on generic
# evidence (P_gain, RE_gene, ClinVar pathogenic, ACMG 3-5) but carry
# neither an ST17 T-ALL cohort hit (T1) nor a candidate-gene overlap (T2).
# For that tier, require consensus support: keep only records whose CALLER
# field indicates MINDA (Severus+SAVANA joint call), drop Severus-only
# T3 hits. T1/T2 are unaffected — biological drivers with clear evidence
# survive regardless of caller consensus, matching the pass2 filter's
# clause A rescue logic.
# CALLER field values seen in pipeline: "SEVERUS", "SAVANA", "MINDA".
# We match "MINDA" substring-inclusively (grepl) so future extensions like
# "MINDA;SEVERUS" or a comma-joined multi-call marker still pass.
#
# Defensive ST17 rescue: keep any record with a populated ST17_hit even
# if it lands in T3 under a future re-tiering (currently T3 excludes ST17
# hits by definition — they'd be T1 — but this guard makes the filter
# resilient to changes in assign_driver_tier() so a T-ALL recurrent hit
# is never silently dropped by the caller-consensus gate).
if ("CALLER" %in% names(pass2_clean)) {
  st17_present <- if ("ST17_hit" %in% names(pass2_clean)) nzchar(pass2_clean$ST17_hit)
                  else rep(FALSE, nrow(pass2_clean))
  pass2_clean <- pass2_clean |>
    filter(!(driver_tier == 3L & !grepl("MINDA", CALLER, fixed = TRUE) & !st17_present))
}

# Dedup Gene_name so the cohort TSV holds "SYMBOL_A; SYMBOL_B" instead of
# AnnotSV's raw full-mode-`|`-joined + split-mode-`,`-repeated concatenation
# (e.g. "LOC1|LOC2|LOC3,LOC1,LOC2,LOC3"). Idempotent — already-clean strings
# round-trip. Keeps Gene_name usable as a stable output-CSV column while
# Gene_name_display remains the DT-friendly capped variant.
if ("Gene_name" %in% names(pass2_clean)) {
  pass2_clean <- pass2_clean |>
    mutate(Gene_name = vapply(Gene_name, dedup_annot_multi,
                              character(1), USE.NAMES = FALSE))
}

# Dedup CytoBand across full/split annotation modes. AnnotSV emits one
# cytoband per gene under split-mode, so multi-gene SVs produce runs like
# "q21.1-q21.3,q21.1,q21.1,q21.2,q21.2,q21.2,...". First dedup to the
# unique set; then collapse to a single "<min>-<max>" encompassing range
# via collapse_cytoband_range() so large SVs don't produce a 30+ band
# semicolon salad in the DT (matches the CNA-side treatment).
if ("CytoBand" %in% names(pass2_clean)) {
  pass2_clean <- pass2_clean |>
    mutate(
      CytoBand = vapply(CytoBand, dedup_annot_multi,
                        character(1), USE.NAMES = FALSE),
      CytoBand = vapply(CytoBand, collapse_cytoband_range,
                        character(1), USE.NAMES = FALSE)
    )
}

# Cap the Gene_name display at 20 genes; append "(n=X genes)" annotation
# when capped so the reader knows the underlying multiplicity. Uses AnnotSV's
# Gene_count column as the authoritative gene tally.
pass2_clean <- pass2_clean |>
  mutate(Gene_name_display = pmap_chr(
    list(Gene_name, Gene_count),
    function(g, n) summarize_gene_list(g, n, cap = 20L)
  ))

# CytoBand display: prepend "chr<CHROM>" to the (already-deduped) band label
# so a clinician reads "chr9p21" instead of a bare "p21". Requires both CHROM
# and CytoBand present; leaves cells blank when either is missing.
if (all(c("CHROM", "CytoBand") %in% names(pass2_clean))) {
  pass2_clean <- pass2_clean |>
    mutate(CytoBand_display = ifelse(
      !is.na(CytoBand) & nzchar(CytoBand) & CytoBand != ".",
      paste0(ifelse(startsWith(as.character(CHROM), "chr"), as.character(CHROM),
                    paste0("chr", CHROM)),
             CytoBand),
      ""
    ))
}

# Optional EnsDb-based BND breakend gene annotation. Skips silently when
# params$ensembl_db is unset OR the file doesn't exist. Handles contigs
# that aren't in the EnsDb (chrUn_*, alt scaffolds, decoys) by filling
# `"unknown_contig:pos"` for the missing side instead of crashing the
# whole join. Vectorized over ALL breakpoints at once (no rowwise) so cost is
# O(N log M) not O(N * M): one `plyranges::join_overlap_inner` for breakpoints
# inside a gene, then one `GenomicRanges::distanceToNearest` for the remainder.
# Only trigger the (potentially slow) Bioc install when the EnsDb SQLite is
# actually staged — otherwise the annotation block would skip anyway.
if (!is.null(params$ensembl_db) && nzchar(params$ensembl_db) &&
    file.exists(params$ensembl_db)) {
  install_bioc_if_missing("ensembldb")
  install_bioc_if_missing("plyranges")
}

edb_annot_available <- !is.null(params$ensembl_db) &&
                        nzchar(params$ensembl_db) &&
                        file.exists(params$ensembl_db) &&
                        requireNamespace("ensembldb", quietly = TRUE) &&
                        requireNamespace("plyranges", quietly = TRUE)

if (edb_annot_available && nrow(pass2_clean) > 0 &&
    all(c("breakpoint", "SVTYPE", "ID") %in% names(pass2_clean))) {

  # Load EnsDb once. GENOME-wide EnsDb includes primary chromosomes 1-22/X/Y/MT.
  edb <- ensembldb::EnsDb(params$ensembl_db)
  # `canonical_transcript` and `gene_biotype` are needed beyond the gene label:
  # the ProteinPaint svfusion export needs an isoform accession per breakend, and
  # biotype is the tie-breaker below. ensembldb does not always surface every
  # underlying gene-table column, so degrade gracefully to gene_name only.
  ens_genes_gr <- tryCatch(
    ensembldb::genes(edb, columns = c("gene_id", "gene_name",
                                      "canonical_transcript", "gene_biotype")),
    error = function(e) {
      warning("EnsDb: canonical_transcript/gene_biotype unavailable (",
              conditionMessage(e), "); falling back to gene_name only.")
      ensembldb::genes(edb, columns = c("gene_id", "gene_name"))
    })
  for (cc in c("canonical_transcript", "gene_biotype")) {
    if (!cc %in% names(S4Vectors::mcols(ens_genes_gr)))
      S4Vectors::mcols(ens_genes_gr)[[cc]] <- NA_character_
  }
  # The overlap/nearest joins keep the QUERY strand ('*' from bp_in_gr), not the
  # subject's, so the gene's own strand must travel as a metadata column.
  ens_genes_gr$gene_strand <- as.character(BiocGenerics::strand(ens_genes_gr))
  # Some EnsDb rows have empty gene_name — fall back to gene_id so the
  # annotation always has a non-empty label.
  ens_genes_gr$gene_name <- ifelse(
    is.na(ens_genes_gr$gene_name) | !nzchar(ens_genes_gr$gene_name),
    ens_genes_gr$gene_id, ens_genes_gr$gene_name
  )
  ens_seqs <- as.character(GenomeInfoDb::seqlevels(ens_genes_gr))

  # Vectorized breakpoint parser — always emits TWO rows per record
  # (left + right anchor) regardless of SVTYPE. Downstream `pivot_wider`
  # produces `gene_name_left`, `distance_left`, `gene_name_right`,
  # `distance_right` columns for every record. Handles three breakpoint
  # shapes emitted by `compute_breakpoint()`:
  #   1. Interchromosomal BND        chrA:posA|chrB:posB
  #   2. Non-BND range (DEL/DUP/INV) chr:start-end
  #   3. Single-point (INS, or END==POS) chr:pos
  # In case (3) left == right (same position). Case (1) matches BND
  # SVTYPE; case (2) matches DEL/DUP/INV; case (3) matches INS or any
  # SVTYPE where END was missing / equal to POS.
  #
  # `distance` is 0 when the anchor is INSIDE the gene (gene_source == "overlap"),
  # positive if the anchor is outside (upstream of the left anchor =
  # negative offset from gene start / downstream of the right anchor =
  # positive offset from gene end). Kept as the raw absolute distance
  # in bp; the direction is implicit from which anchor (left / right).
  bp_split <- function(bp, sv, rid) {
    if (is.na(bp) || !nzchar(bp)) return(NULL)
    sides <- str_split(bp, "\\|")[[1]]

    # Case 1: interchromosomal BND (two |-separated chr:pos anchors)
    if (length(sides) == 2) {
      m1 <- str_match(sides[1], "^(chr[^:]+):([0-9]+)$")
      m2 <- str_match(sides[2], "^(chr[^:]+):([0-9]+)$")
      if (any(is.na(m1[1, ]) | is.na(m2[1, ]))) return(NULL)
      return(data.frame(
        ID       = c(rid, rid),
        side     = c("left", "right"),
        seqnames = c(sub("^chr", "", m1[1, 2]),
                     sub("^chr", "", m2[1, 2])),
        start    = as.integer(c(m1[1, 3], m2[1, 3])),
        stringsAsFactors = FALSE
      ))
    }

    # Case 2: non-BND range (DEL / DUP / INV) — one chr, two positions
    m <- str_match(sides[1], "^(chr[^:]+):([0-9]+)-([0-9]+)$")
    if (!is.na(m[1, 1])) {
      return(data.frame(
        ID       = c(rid, rid),
        side     = c("left", "right"),
        seqnames = sub("^chr", "", m[1, 2]),
        start    = as.integer(c(m[1, 3], m[1, 4])),
        stringsAsFactors = FALSE
      ))
    }

    # Case 3: single-point (INS with END==POS, or SVTYPE where END is
    # missing) — left == right. Distance columns will match on both sides.
    m <- str_match(sides[1], "^(chr[^:]+):([0-9]+)$")
    if (!is.na(m[1, 1])) {
      p <- as.integer(m[1, 3])
      return(data.frame(
        ID       = c(rid, rid),
        side     = c("left", "right"),
        seqnames = sub("^chr", "", m[1, 2]),
        start    = c(p, p),
        stringsAsFactors = FALSE
      ))
    }

    NULL
  }

  bp_df <- purrr::pmap(
    list(pass2_clean$breakpoint, pass2_clean$SVTYPE, pass2_clean$ID),
    bp_split
  ) |> purrr::compact() |> dplyr::bind_rows()

  if (nrow(bp_df) > 0) {
    bp_df$end <- bp_df$start + 1L
    # Split into in-EnsDb vs out-of-EnsDb before the join to avoid
    # 0-hit crashes for contigs like `chrUn_GL000219v1`.
    bp_in  <- bp_df |> dplyr::filter(seqnames %in% ens_seqs)
    bp_out <- bp_df |> dplyr::filter(!(seqnames %in% ens_seqs)) |>
      dplyr::mutate(gene_name = "unknown_contig", distance = NA_integer_)

    if (nrow(bp_in) > 0) {
      bp_in_gr <- plyranges::as_granges(bp_in)

      # TWO-STAGE ASSIGNMENT. `join_nearest` alone is NOT usable here: plyranges
      # documents it as returning an ARBITRARY nearest neighbour, and it offers no
      # multi-mapping option, so when a breakpoint sits inside several features it
      # discards the competitors BEFORE any tie-break can see them. The ordering
      # below was never wrong — it was being handed a one-row candidate set.
      #
      # Observed 2026-09-17, re-confirmed against EnsDb v110 on 2026-10-02:
      # chr9:131151786 lies inside BOTH NUP214 (ENSG00000126883, protein_coding,
      # chr9:131125573-131234663) and an unnamed lncRNA (ENSG00000236986,
      # chr9:131132852-131167505). join_nearest returned only the lncRNA, so
      # SET::NUP214 — the canonical T-ALL fusion from del(9)(q34.11q34.13) — was
      # emitted as SET::ENSG00000236986.
      #
      # Stage 1 uses join_overlap_inner (== find_overlaps), which emits one row per
      # (breakpoint, gene) pair, so every containing feature competes.
      # Stage 2 handles breakpoints in gene deserts, where there is nothing to
      # overlap, via distanceToNearest(select = "all") — the only API here that
      # returns ALL equidistant genes rather than one arbitrary pick.
      #
      # LRG_gene features are dropped at source. EnsDb carries LRG duplicates of
      # real genes (NUP214 is both ENSG00000126883 and LRG_1383 at near-identical
      # coordinates) whose canonical_transcript is an LRG accession such as
      # `LRG_482t1`. Those are neither RefSeq nor Ensembl, so they are rejected by
      # the ProteinPaint svfusion spec and degrade to a featureless box.
      ens_sel <- ens_genes_gr[
        is.na(ens_genes_gr$gene_biotype) | ens_genes_gr$gene_biotype != "LRG_gene"]
      ens_sel$gene_width <- BiocGenerics::width(ens_sel)
      ens_sel <- ens_sel[, c("gene_name", "canonical_transcript",
                             "gene_biotype", "gene_strand", "gene_width")]

      # Keep one row per (ID, side): downstream pivot_wider must produce atomic
      # character columns, not list-columns (which break coalesce in mlst_bnd).
      # Order: nearest, then protein_coding ahead of other biotypes, then an
      # official symbol ahead of a bare ENSG id (line ~1523 has already coalesced
      # an empty gene_name to gene_id, so the ENSG test does fire), then the
      # tightest gene interval LAST. Width must rank last and not earlier: the
      # lncRNA above is 34,653 bp against NUP214's 109,090 bp, so a width-first
      # rule actively prefers the wrong gene.
      .pick_one <- function(df) {
        df |>
          dplyr::arrange(ID, side, distance,
                         dplyr::coalesce(gene_biotype, "zzz") != "protein_coding",
                         startsWith(gene_name, "ENSG"),
                         gene_width) |>
          dplyr::distinct(ID, side, .keep_all = TRUE)
      }

      # ── Stage 1: every gene CONTAINING the breakpoint ──────────────────────
      ov <- plyranges::join_overlap_inner(bp_in_gr, ens_sel) |>
        tibble::as_tibble() |>
        dplyr::mutate(distance = 0L, gene_source = "overlap",
                      nearest_genes = NA_character_) |>
        dplyr::select(ID, side, gene_name, canonical_transcript, gene_biotype,
                      gene_strand, distance, gene_width, gene_source,
                      nearest_genes)

      # ── Stage 2: breakpoints with no containing gene -> ALL nearest ────────
      .done <- paste(ov$ID, ov$side)
      rest  <- bp_in_gr[!paste(bp_in_gr$ID, bp_in_gr$side) %in% .done]

      near <- ov[0, ]
      if (length(rest) > 0) {
        h <- GenomicRanges::distanceToNearest(rest, ens_sel, select = "all",
                                              ignore.strand = TRUE)
        if (length(h) > 0) {
          qi <- S4Vectors::queryHits(h); si <- S4Vectors::subjectHits(h)
          near <- tibble::tibble(
            ID                   = rest$ID[qi],
            side                 = rest$side[qi],
            gene_name            = ens_sel$gene_name[si],
            canonical_transcript = ens_sel$canonical_transcript[si],
            gene_biotype         = ens_sel$gene_biotype[si],
            gene_strand          = ens_sel$gene_strand[si],
            distance             = S4Vectors::mcols(h)$distance,
            gene_width           = ens_sel$gene_width[si],
            gene_source          = "nearest") |>
            # Report EVERY nearest gene, not just the one that wins the
            # tie-break: an intergenic breakpoint is only interpretable with its
            # flanking context. The primary columns stay atomic for pivot_wider;
            # this collapsed string is the full picture for the report.
            dplyr::group_by(ID, side) |>
            dplyr::mutate(nearest_genes = paste0(
              paste(sort(unique(gene_name)), collapse = ", "),
              " (", dplyr::first(distance), " bp)")) |>
            dplyr::ungroup()
        }
      }

      hits <- .pick_one(dplyr::bind_rows(ov, near)) |>
        dplyr::select(-gene_width)
    } else hits <- tibble::tibble(ID = character(), side = character(),
                                    gene_name = character(),
                                    canonical_transcript = character(),
                                    gene_biotype = character(),
                                    gene_strand = character(),
                                    gene_source = character(),
                                    nearest_genes = character(),
                                    distance = integer())

    bp_out_hits <- bp_out |>
      dplyr::mutate(canonical_transcript = NA_character_,
                    gene_biotype         = NA_character_,
                    gene_strand          = NA_character_,
                    gene_source          = "unknown_contig",
                    nearest_genes        = NA_character_) |>
      dplyr::select(ID, side, gene_name, canonical_transcript,
                    gene_biotype, gene_strand, distance, gene_source,
                    nearest_genes)
    all_hits <- dplyr::bind_rows(hits, bp_out_hits)

    # Every record now has a `left` and a `right` row (see bp_split
    # docs). Pivot to per-side gene / distance columns so every record
    # ends up with 4 new columns: gene_name_left, distance_left,
    # gene_name_right, distance_right.
    lr <- all_hits |>
      dplyr::filter(side %in% c("left", "right")) |>
      tidyr::pivot_wider(id_cols     = ID,
                         names_from  = side,
                         values_from = c(gene_name, canonical_transcript,
                                         gene_strand, distance, gene_source,
                                         nearest_genes),
                         names_glue  = "{.value}_{side}")

    pass2_clean <- pass2_clean |>
      dplyr::left_join(lr, by = "ID") |>
      dplyr::mutate(
        # For BNDs the pair reads as a fusion (matches ST17_hit shape:
        # `GENE_A::GENE_B`). For non-BND records the two anchors are on
        # the same chromosome so use an ASCII arrow to show the local range.
        # ENSG identifiers are preserved when EnsDb had no HGNC symbol —
        # they carry biological meaning (pseudogene, lincRNA, novel locus)
        # and should not be silently masked. `unknown_contig` from the
        # bp_out fallback is passed through unchanged.

        edb_gene_pair = dplyr::case_when(
          SVTYPE == "BND" & !is.na(gene_name_left) & !is.na(gene_name_right) ~
            paste0(gene_name_left, "::", gene_name_right),
          !is.na(gene_name_left) & !is.na(gene_name_right) ~
            paste0(gene_name_left, " <-> ", gene_name_right),
          TRUE ~ NA_character_
        ),
        # Compact distance summary — "L: 0 bp; R: 12503 bp" reads at a
        # glance; distance == 0 means the anchor lies inside the gene.
        edb_gene_distance = dplyr::case_when(
          !is.na(distance_left) & !is.na(distance_right) ~
            sprintf("L: %s bp; R: %s bp", distance_left, distance_right),
          TRUE ~ NA_character_
        )
      )
  }
}

# has_mate — True when both _1 and _2 mates of a BND survive pass2.
# Simple: strip trailing `_1` / `_2` and count records per stripped ID.
# Non-BND records (no _1/_2 suffix) get has_mate = FALSE (mate concept
# doesn't apply). Useful for sort / filter in the DT: T1 BND records
# with has_mate=TRUE are the highest-confidence fusion calls.
pass2_clean <- pass2_clean |>
  group_by(ID_SHORT = str_remove(as.character(ID), "_[12]$")) |>
  mutate(has_mate = (SVTYPE == "BND" & dplyr::n() == 2L)) |>
  ungroup() |>
  select(-ID_SHORT)

# Attach full sample_metadata (all cohort fields) so the cohort TSV carries
# clinical context — sex, age, EGIL, ETP, karyotype, driver fusions, ITD, etc.
# No-op when the metadata CSV wasn't supplied.
pass2_clean <- merge_sample_metadata(pass2_clean)

# Emit the cohort-level SV pass2 TSV (full schema, RE_gene retained,
# Closest_left/Closest_right + new driver_tier / tier_score / ST17_hit /
# Gene_name_display + CytoBand_display + has_mate columns included). The
# SV_REPORT_INDEX module declares `optional: true` on tsv_sv so the write
# being suppressed by the chunk-level guard is OK when pass2 is empty.
readr::write_tsv(pass2_clean, "cohort_sv_pass2.tsv", na = "NA")
```

```{r load-cna-pass2}
# CNA pass2 TSVs live under `cna_pass2/` to avoid colliding with the SV
# pass2 glob above — both file shapes end in `.filtered.pass2.tsv`.
# CN-specific columns: TCN/CN1/CN2/CNQ1/CNQ2/COV1/COV2 (vs. SV's VAF/hVAF/DR/DV)

cna_pass2_files <- list.files(
  path       = "cna_pass2",
  pattern    = "\\.filtered\\.pass2\\.tsv$",
  full.names = TRUE
)

cna_pass2_long <- if (length(cna_pass2_files) > 0) {
  map_dfr(cna_pass2_files, ~ read_tsv(.x, col_types = cols(.default = "c"),
                                      show_col_types = FALSE))
} else {
  tibble()
}

if (nrow(cna_pass2_long) > 0 && !"SAMPLE" %in% names(cna_pass2_long)) {
  cna_pass2_long <- cna_pass2_long |> mutate(SAMPLE = "unknown")
}

# Derive a per-record CNA type from ALT (`<DUP>` -> "DUP") because SVTYPE in
# Wakhan VCFs is uniformly "CNV" and doesn't differentiate gain vs. loss.
if (nrow(cna_pass2_long) > 0) {
  cna_pass2_long <- cna_pass2_long |>
    mutate(
      CNA_TYPE         = str_remove_all(as.character(ALT), "[<>]"),
      ACMG_class_clean = acmg_factor(ACMG_class)
    )
}
```

```{r cna-pass2-clean}
#| eval: !expr nrow(cna_pass2_long) > 0


# write the cohort TSV 
cna_pass2_long <- cna_pass2_long |>
  mutate(across(where(is.character), clean_tuple))

# Union candidate-gene detection across Gene_name + Closest_left + Closest_right
if (length(candidate_set) > 0 && "Gene_name" %in% names(cna_pass2_long)) {
  closest_l <- if ("Closest_left"  %in% names(cna_pass2_long)) cna_pass2_long$Closest_left  else rep("", nrow(cna_pass2_long))
  closest_r <- if ("Closest_right" %in% names(cna_pass2_long)) cna_pass2_long$Closest_right else rep("", nrow(cna_pass2_long))
  cna_pass2_long <- cna_pass2_long |>
    mutate(Candidate_genes = pmap_chr(
      list(Gene_name, closest_l, closest_r),
      function(gn, cl, cr) compute_candidate_genes(gn, cl, cr, candidate_set)
    ))
} else {
  cna_pass2_long$Candidate_genes <- ""
}

cna_pass2_clean <- cna_pass2_long |>
  mutate(ACMG_class = acmg_clean(ACMG_class)) |>
  select(-any_of("Annotation_mode")) |>
  mutate(Gene_name_display = pmap_chr(
    list(Gene_name, Gene_count),
    function(g, n) summarize_gene_list(g, n, cap = 20L)
  ))

# breakpoint column: "chr21:13000002-31877979" for Wakhan CNA segments. Wakhan records are always intra-chromosomal
if (all(c("CHROM", "POS", "END", "ALT") %in% names(cna_pass2_clean))) {
  svtype_col <- if ("SVTYPE" %in% names(cna_pass2_clean)) cna_pass2_clean$SVTYPE else rep("CNV", nrow(cna_pass2_clean))
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(breakpoint = compute_breakpoint(CHROM, POS, END, ALT, svtype_col))
}

# Parse WAKHAN_TYPE (GAIN / LOSS) from WAKHAN_ID's second colon-separated
# WAKHAN_ID format: "wakhan:GAIN:chr21:13000002-31877979".
# Falls back to the ALT-derived CNA_TYPE if WAKHAN_ID is missing 
if ("WAKHAN_ID" %in% names(cna_pass2_clean)) {
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(WAKHAN_TYPE = case_when(
      !is.na(WAKHAN_ID) & nzchar(WAKHAN_ID) ~ str_split_fixed(WAKHAN_ID, ":", 3)[, 2],
      CNA_TYPE == "DUP" ~ "GAIN",
      CNA_TYPE == "DEL" ~ "LOSS",
      TRUE ~ ""
    ))
} else {
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(WAKHAN_TYPE = case_when(
      CNA_TYPE == "DUP" ~ "GAIN",
      CNA_TYPE == "DEL" ~ "LOSS",
      TRUE ~ ""
    ))
}

# Summarize P_gain_source / P_loss_source / P_ins_source 
# DT shows the count and first-3 exemplars of each source
p_source_cols <- intersect(
  c("P_gain_source", "P_loss_source", "P_ins_source"),
  names(cna_pass2_clean)
)
if (length(p_source_cols) > 0) {
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(across(all_of(p_source_cols),
                  ~ vapply(.x, summarize_p_source, character(1),
                           USE.NAMES = FALSE)))
}

#  ST19 name/score cells match the CNA type suffix (AMP/DEL/GAIN/LOSS) 
cna_pass2_clean <- cna_pass2_clean |>
  mutate(across(
    matches("^(name|score)_ST19_Alterations\\.CNV\\.Recurrent_(AMP|DEL|GAIN|LOSS)$"),
    function(col_vals) {
      suffix <- sub("^(name|score)_ST19_Alterations\\.CNV\\.Recurrent_", "",
                    cur_column())
      is_gain_like <- (WAKHAN_TYPE == "GAIN" | CNA_TYPE == "DUP")
      is_loss_like <- (WAKHAN_TYPE == "LOSS" | CNA_TYPE == "DEL")
      keep <- (is_gain_like & suffix %in% c("AMP", "GAIN")) |
              (is_loss_like & suffix %in% c("DEL", "LOSS"))
      ifelse(keep, as.character(col_vals), "")
    }
  ))

# CNA driver-priority tier 
cna_tier_tbl <- assign_driver_tier_cna(
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_AMP"]],
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_DEL"]],
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_GAIN"]],
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_LOSS"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_AMP"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_DEL"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_GAIN"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_LOSS"]],
  cna_pass2_clean$Candidate_genes,
  cna_pass2_clean$AnnotSV_ranking_score
)
cna_pass2_clean <- bind_cols(cna_pass2_clean, cna_tier_tbl)

# Consolidate the 4 SVTYPE-matched ST19 name / score columns into a single
# top-3 hit summary — mirrors the ST17 pattern on the SV side (see the
# `pass2-clean` chunk). Rationale: the 8 raw ST19 columns are noisy (per-
# BED cells padded with `.,.,.` and misaligned between name and score
# order), which makes the DT hard to read. This block:
#   * Concatenates the 4 SVTYPE-matched `name_ST19_*` cells and the 4
#     `score_ST19_*` cells into pipe-joined raw strings (the earlier
#     SVTYPE-mismatch blanking has already left only the correct suffix
#     populated per record — GAIN+AMP for GAIN-like, DEL+LOSS for LOSS-like).
#   * Runs the same `parse_st17_pair()` helper (name-agnostic despite its
#     ST17 naming): splits by `|`, pairs by index, dedups by name,
#     orders by score desc, truncates to top-3.
#   * Emits three consolidated columns: `ST19_name`, `ST19_score`, and
#     `ST19_hit` (which renders "hit1 [s1]; hit2 [s2]; hit3 [s3]" for the DT).
# The raw 8 columns are retained in cna_pass2_clean for TSV export (they
# still land in cohort_cna_pass2.tsv) but dropped from the DT display below.
join_pipe_cells <- function(...) {
  parts <- as.character(c(...))
  parts <- parts[!is.na(parts) & nzchar(parts) & parts != "."]
  if (length(parts) == 0) return("")
  paste(parts, collapse = "|")
}

st19_name_raw <- mapply(
  join_pipe_cells,
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_AMP"]],
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_DEL"]],
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_GAIN"]],
  cna_pass2_clean[["name_ST19_Alterations.CNV.Recurrent_LOSS"]],
  USE.NAMES = FALSE
)
st19_score_raw <- mapply(
  join_pipe_cells,
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_AMP"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_DEL"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_GAIN"]],
  cna_pass2_clean[["score_ST19_Alterations.CNV.Recurrent_LOSS"]],
  USE.NAMES = FALSE
)

st19_pairs <- purrr::map2(st19_name_raw, st19_score_raw,
                          ~ parse_st17_pair(.x, .y, top_n = 3L))

cna_pass2_clean <- cna_pass2_clean |>
  mutate(
    ST19_name  = vapply(st19_pairs, `[[`, character(1), "names"),
    ST19_score = vapply(st19_pairs, `[[`, character(1), "scores"),
    ST19_hit   = vapply(seq_along(ST19_name), function(i) {
      nm <- ST19_name[i]; sc <- ST19_score[i]
      if (!nzchar(nm)) return("")
      nms <- split_pipe_outside_parens(nm)
      scs <- if (nzchar(sc)) str_split(sc, "\\|")[[1]] else rep("", length(nms))
      length(scs) <- length(nms)
      scs[is.na(scs)] <- ""
      pieces <- vapply(seq_along(nms), function(j) {
        if (nzchar(scs[j])) paste0(nms[j], " [", scs[j], "]") else nms[j]
      }, character(1))
      paste(pieces, collapse = "; ")
    }, character(1))
  )

# Same cleanup pass as SV: strip AnnotSV `full,split,split,...` padding
# from every character column, then dedup Gene_name / CytoBand across
# full/split modes, then compute CytoBand_display = "chr<CHROM><band>".
cna_pass2_clean <- cna_pass2_clean |>
  mutate(across(where(is.character), \(v) vapply(v, clean_annot_padding, character(1), USE.NAMES = FALSE)))

if ("Gene_name" %in% names(cna_pass2_clean)) {
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(Gene_name = vapply(Gene_name, dedup_annot_multi,
                              character(1), USE.NAMES = FALSE))
}

if ("CytoBand" %in% names(cna_pass2_clean)) {
  # Two-step cleanup: dedup_annot_multi() first turns the raw AnnotSV
  # full+split cell into a semicolon-joined unique-list; then
  # collapse_cytoband_range() folds that list to a single "<min>-<max>"
  # range so the DT shows "q23.31-q26.3" instead of 30+ semicolon-joined
  # sub-bands. Large-segment CNAs (very common on chr9p21, chr10q, etc.)
  # were the pain point that motivated this.
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(
      CytoBand = vapply(CytoBand,       dedup_annot_multi,
                        character(1), USE.NAMES = FALSE),
      CytoBand = vapply(CytoBand,       collapse_cytoband_range,
                        character(1), USE.NAMES = FALSE)
    )
}

if (all(c("CHROM", "CytoBand") %in% names(cna_pass2_clean))) {
  cna_pass2_clean <- cna_pass2_clean |>
    mutate(CytoBand_display = ifelse(
      !is.na(CytoBand) & nzchar(CytoBand) & CytoBand != ".",
      paste0(ifelse(startsWith(as.character(CHROM), "chr"), as.character(CHROM),
                    paste0("chr", CHROM)),
             CytoBand),
      ""
    ))
}

cna_pass2_clean <- merge_sample_metadata(cna_pass2_clean)

readr::write_tsv(cna_pass2_clean, "cohort_cna_pass2.tsv", na = "NA")
```

```{r load-snv}
# FILTER_SNV TSVs are staged under snv/ by the module's stageAs directive.
snv_files <- list.files(path = "snv", pattern = "\\.snv\\.filtered\\.tsv$",
                         full.names = TRUE)

snv_long <- if (length(snv_files) > 0) {
  map_dfr(snv_files, ~ read_tsv(.x, col_types = cols(.default = "c"),
                                show_col_types = FALSE))
} else {
  tibble()
}

if (nrow(snv_long) > 0 && !"SAMPLE" %in% names(snv_long)) {
  snv_long <- snv_long |> mutate(SAMPLE = "unknown")
}

# Coerce numerics needed by plots / filters
if (nrow(snv_long) > 0) {
  snv_long <- snv_long |>
    mutate(
      gnomADg_AF      = suppressWarnings(as.numeric(gnomADg_AF)),
      CLAIRSTO_VAF    = suppressWarnings(as.numeric(CLAIRSTO_VAF)),
      DEEPSOMATIC_VAF = suppressWarnings(as.numeric(DEEPSOMATIC_VAF))
    )
}
```

```{r snv-clean}
#| eval: !expr nrow(snv_long) > 0

# Candidate_genes: SYMBOL i∩ candidate gene panel
if (length(candidate_set) > 0 && "SYMBOL" %in% names(snv_long)) {
  snv_long <- snv_long |>
    mutate(Candidate_genes = ifelse(SYMBOL %in% candidate_set, SYMBOL, ""))
} else {
  snv_long$Candidate_genes <- ""
}

# Split the CIVIC INFO field into named columns for downstream review. Skips
# silently when the annotate_civic_snv step wasn't staged (no CIVIC column).
# The seven CIViC_* columns land immediately after the raw CIVIC column so
# reviewers can compare source and parsed values side-by-side.
if ("CIVIC" %in% names(snv_long)) {
  civic_parsed <- purrr::map(snv_long$CIVIC, parse_civic_field)
  civic_tbl <- tibble::tibble(
    CIViC_Consequence           = vapply(civic_parsed, `[[`, character(1), "CIViC_Consequence"),
    CIViC_SYMBOL                = vapply(civic_parsed, `[[`, character(1), "CIViC_SYMBOL"),
    CIViC_Variant_ID            = vapply(civic_parsed, `[[`, character(1), "CIViC_Variant_ID"),
    CIViC_Variant_URL           = vapply(civic_parsed, `[[`, character(1), "CIViC_Variant_URL"),
    CIViC_Entity_Variant_Origin = vapply(civic_parsed, `[[`, character(1), "CIViC_Entity_Variant_Origin"),
    CIViC_Entity_Therapies      = vapply(civic_parsed, `[[`, character(1), "CIViC_Entity_Therapies"),
    CIViC_Assertion_ACMG_Codes  = vapply(civic_parsed, `[[`, character(1), "CIViC_Assertion_ACMG_Codes")
  )
  snv_long <- dplyr::bind_cols(snv_long, civic_tbl)
}

# SNV driver-priority tier
snv_tier_tbl <- assign_driver_tier_snv(
  onc             = if ("ONC"             %in% names(snv_long)) snv_long$ONC             else rep(NA_character_, nrow(snv_long)),
  clnsig          = if ("CLNSIG"          %in% names(snv_long)) snv_long$CLNSIG          else rep(NA_character_, nrow(snv_long)),
  candidate_genes = snv_long$Candidate_genes,
  clairsto_vaf    = if ("CLAIRSTO_VAF"    %in% names(snv_long)) snv_long$CLAIRSTO_VAF    else rep(NA_character_, nrow(snv_long)),
  deepsomatic_vaf = if ("DEEPSOMATIC_VAF" %in% names(snv_long)) snv_long$DEEPSOMATIC_VAF else rep(NA_character_, nrow(snv_long))
)
snv_long <- bind_cols(snv_long, snv_tier_tbl)

snv_long <- merge_sample_metadata(snv_long)

# Emit cohort TSV (full schema, including columns not shown in DT). Optional
# output of the SV_REPORT_INDEX module.
readr::write_tsv(snv_long, "cohort_snv.tsv", na = "NA")
```

```{r disco-plot-prep}
#| eval: !expr isTRUE(params$disco_enable)

# ProteinPaint DISCO expects an `mlst` (mutation list) array with per-variant objects. Data-type (`dt`) codes: 1 = SNV point mutation, 2 = SV / fusion, 4 = CNV segment (log2 copy-number ratio as `value`).


#  SNV mlst 
disco_snv_class <- function(consequence) {
  dplyr::case_when(
    is.na(consequence) | !nzchar(consequence)                                             ~ "M",
    grepl("stop_gained|stop_lost|start_lost|nonsense", consequence, ignore.case = TRUE)   ~ "N",
    grepl("frameshift",                                consequence, ignore.case = TRUE)   ~ "F",
    grepl("splice",                                    consequence, ignore.case = TRUE)   ~ "P",
    grepl("inframe_deletion",                          consequence, ignore.case = TRUE)   ~ "D",
    grepl("inframe_insertion",                         consequence, ignore.case = TRUE)   ~ "I",
    grepl("missense",                                  consequence, ignore.case = TRUE)   ~ "M",
    grepl("synonymous",                                consequence, ignore.case = TRUE)   ~ "S",
    TRUE                                                                                  ~ "M"
  )
}

mlst_snv <- if (exists("snv_long") && nrow(snv_long) > 0 &&
                "driver_tier" %in% names(snv_long)) {
  snv_long |>
    dplyr::filter(driver_tier == 1L) |>
    dplyr::transmute(
      dt       = 1L,
      chr      = if (all(startsWith(as.character(CHROM), "chr"))) as.character(CHROM)
                 else paste0("chr", CHROM),
      position = suppressWarnings(as.integer(POS)),
      ref      = REF,
      alt      = ALT,
      gene     = dplyr::coalesce(SYMBOL, "intergenic"),
      isoform  = dplyr::coalesce(Feature, "unknown"),
      mname    = dplyr::coalesce(HGVSp, HGVSc, ""),
      class    = disco_snv_class(Consequence),
      sample   = if ("sample_id" %in% names(snv_long))
                   dplyr::coalesce(sample_id, SAMPLE) else SAMPLE
    ) |>
    dplyr::filter(!is.na(position))
} else tibble::tibble()

#  CNA mlst 
mlst_cna <- if (exists("cna_pass2_clean") && nrow(cna_pass2_clean) > 0 &&
                "driver_tier" %in% names(cna_pass2_clean)) {
  cna_pass2_clean |>
    dplyr::filter(driver_tier == 1L) |>
    dplyr::transmute(
      dt     = 4L,
      chr    = if (all(startsWith(as.character(CHROM), "chr"))) as.character(CHROM)
               else paste0("chr", CHROM),
      start  = suppressWarnings(as.integer(POS)),
      stop   = suppressWarnings(as.integer(END)),
      value  = suppressWarnings(as.numeric(TCN)),
      # value = suppressWarnings(log2(pmax(as.numeric(TCN), 0.25) / 2))
      sample = if ("sample_id" %in% names(cna_pass2_clean))
                 dplyr::coalesce(sample_id, SAMPLE) else SAMPLE
    ) |>
    dplyr::filter(!is.na(start), !is.na(stop), !is.na(value))
} else tibble::tibble()

#  SV BND mlst
# Partner chr + pos parsed from ALT bracket notation via `parse_bnd_partner()`.
# `geneA` / `geneB` (record-side and partner-side gene) come from the bp_split
# EnsDb annotation earlier in the pipeline: `gene_name_left` is the anchor at
# CHROM:POS and `gene_name_right` is the partner anchor. Fall back to
# AnnotSV/VEP-derived `Gene_name` / `Closest_left` for `geneA` and "unknown"
# for `geneB` when the EnsDb block didn't run (e.g. rocker/verse image without
# ensembldb + plyranges).
mlst_bnd <- if (exists("pass2_clean") && nrow(pass2_clean) > 0 &&
                "driver_tier" %in% names(pass2_clean)) {
  bnd_t1 <- pass2_clean |>
    dplyr::filter(driver_tier == 1L, SVTYPE == "BND")
  if (nrow(bnd_t1) > 0) {
    partners  <- lapply(bnd_t1$ALT, parse_bnd_partner)
    prefix_chr <- function(c) ifelse(is.na(c) | !nzchar(c) | startsWith(c, "chr"),
                                     c, paste0("chr", c))
    gene_left  <- if ("gene_name_left"  %in% names(bnd_t1)) bnd_t1$gene_name_left
                  else rep(NA_character_, nrow(bnd_t1))
    gene_right <- if ("gene_name_right" %in% names(bnd_t1)) bnd_t1$gene_name_right
                  else rep(NA_character_, nrow(bnd_t1))
    sample_lbl <- if ("sample_id" %in% names(bnd_t1))
                    dplyr::coalesce(bnd_t1$sample_id, bnd_t1$SAMPLE)
                  else bnd_t1$SAMPLE
    dplyr::tibble(
      dt      = 2L,
      chrA    = prefix_chr(as.character(bnd_t1$CHROM)),
      posA    = suppressWarnings(as.integer(bnd_t1$POS)),
      geneA   = dplyr::coalesce(gene_left, bnd_t1$Gene_name,
                                bnd_t1$Closest_left, "intergenic"),
      # parse_bnd_partner strips the "chr" prefix via normalize_chrom() —
      # re-prepend it here since DISCO expects "chrN" format.
      chrB    = prefix_chr(vapply(partners, function(p) as.character(p$chrom),
                                  character(1))),
      posB    = vapply(partners, function(p) suppressWarnings(as.integer(p$pos)),
                       integer(1)),
      geneB   = dplyr::coalesce(gene_right, "unknown"),
      strandA = "+",
      strandB = "-",
      sample  = sample_lbl
    ) |>
      dplyr::filter(!is.na(posA), !is.na(chrB), !is.na(posB))
  } else tibble::tibble()
} else tibble::tibble()

# Combine as a single mlst list — but transpose EACH subset first so per-
# record named lists only carry that dt's own fields. bind_rows() would
# NA-pad missing columns (SNV records would get chrA/posA/geneA=NA, CNA
# records would get chr/position/gene=NA, etc.) and jsonlite serializes
# those as literal `null` which ProteinPaint DISCO rejects as malformed
# records. Per-subset transpose keeps each dt's payload minimal.
df_to_records <- function(df) {
  if (nrow(df) == 0) return(list())
  purrr::transpose(df)
}
mlst_all_list <- c(
  df_to_records(mlst_snv),
  df_to_records(mlst_cna),
  df_to_records(mlst_bnd)
)

# Keep `mlst_all` as a tibble for the status-count chunk below (`sum(dt==1)`
# etc.); the JSON step uses `mlst_all_list` for the record-level output.
mlst_all <- dplyr::bind_rows(mlst_snv, mlst_cna, mlst_bnd)
```

```{r disco-plot-data, include = FALSE}
#| eval: !expr isTRUE(params$disco_enable) && exists("mlst_all") && nrow(mlst_all) > 0

# DATA-ONLY chunk. This chunk produces no rendered output; it exposes a
# handful of variables (disco_mlst_json, safe_cohort_label, pp_host,
# pp_genome, n_cohort_snv/cna/bnd, n_cohort_samples) that the pure-markdown
# runproteinpaint block below then consumes via knitr inline `r ...`
# expressions. Separating data from presentation this way keeps the
# runproteinpaint template inspectable as real HTML/JS with syntax
# highlighting instead of a sprintf-embedded string blob.
#
# Layout produced by this chunk + the markdown below:
#   * ONE cohort DISCO plot rendered inline in the report body (single
#     `<div id="disco_pp_cohort">` mount + runproteinpaint call). Contains
#     the union of Tier-1 records across every sample.
#   * Per-sample DISCO HTMLs written to `protein_paint_out/` (still
#     managed by this chunk via build_disco_html + writeLines) for
#     sample-scoped inspection outside the main report.
# No top-level `disco_plot.html` file is written any more — the report
# itself IS the cohort plot.

pp_host   <- if (!is.null(params$proteinpaint_host)   && nzchar(params$proteinpaint_host))   params$proteinpaint_host   else "http://localhost:3456"
pp_genome <- if (!is.null(params$proteinpaint_genome) && nzchar(params$proteinpaint_genome)) params$proteinpaint_genome else "hg38"

# Sanitize label for safe use as a filename component: letters, digits,
# dot, underscore, hyphen — anything else becomes underscore.
disco_sample_slug <- function(s) {
  s <- as.character(s)
  s[is.na(s) | !nzchar(s)] <- "unknown"
  gsub("[^A-Za-z0-9._-]+", "_", s)
}

# Build the full standalone HTML document for a per-sample DISCO plot.
# Only used for the per-sample files under protein_paint_out/. The cohort
# plot is rendered inline in this document; nothing here is emitted for it.
build_disco_html <- function(mlst_records, ring_label) {
  mlst_records <- lapply(mlst_records, function(rec) rec[names(rec) != "sample"])
  mlst_json <- jsonlite::toJSON(mlst_records, auto_unbox = TRUE, na = "null")
  safe_label <- gsub('"', '\\"', gsub("\\\\", "\\\\\\\\", as.character(ring_label)), fixed = TRUE)
  sprintf('<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>DISCO plot — %s</title>
  <style>
    body { margin: 0; padding: 12px; font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", sans-serif; background: #ffffff; }
    #disco_pp_holder { min-height: 700px; width: 100%%; }
  </style>
  <script src="https://proteinpaint.stjude.org/bin/proteinpaint.js" charset="utf-8"></script>
</head>
<body>
  <div id="disco_pp_holder"></div>
  <script>
    document.addEventListener("DOMContentLoaded", function () {
      if (typeof runproteinpaint !== "function") {
        document.getElementById("disco_pp_holder").innerHTML =
          "<em>ProteinPaint JS bundle failed to load. Verify network access to proteinpaint.stjude.org or host a local mirror.</em>";
        return;
      }
      runproteinpaint({
        host: "%s",
        holder: document.getElementById("disco_pp_holder"),
        noheader: true,
        nobox: true,
        genome: "%s",
        disco: {
          mlst: %s,
          settings: { Disco: { centerText: "%s" } }
        }
      });
    });
  </script>
</body>
</html>
', safe_label, pp_host, pp_genome, mlst_json, safe_label)
}

# Partition records by sample for per-sample HTMLs.
sample_of <- function(rec) {
  v <- rec[["sample"]]
  if (is.null(v) || is.na(v) || !nzchar(as.character(v))) "unknown" else as.character(v)
}
per_sample_labels  <- vapply(mlst_all_list, sample_of, character(1))
per_sample_split   <- split(mlst_all_list, per_sample_labels)
sample_ids         <- names(per_sample_split)
sample_slugs       <- disco_sample_slug(sample_ids)
n_cohort_samples   <- length(sample_ids)

# Braces required: R's parser closes the outer `if (…) expr` before it
# reaches the naked `else` on the next line, so the chained form must
# be `if {…} else if {…} else {…}` when broken across lines.
cohort_label <- if (n_cohort_samples > 1) {
  sprintf("Cohort (n = %d)", n_cohort_samples)
} else if (n_cohort_samples == 1) {
  sample_ids[1]
} else {
  "Cohort"
}

# Per-sample DISCO HTMLs → protein_paint_out/.
dir.create("protein_paint_out", showWarnings = FALSE, recursive = TRUE)
sample_files <- file.path("protein_paint_out",
                          paste0("disco_plot_", sample_slugs, ".html"))
for (i in seq_along(sample_ids)) {
  writeLines(
    build_disco_html(per_sample_split[[i]], sample_ids[i]),
    sample_files[i]
  )
}

# ── Values consumed by the pure-markdown runproteinpaint block below via
# inline `r ...` expressions. Every hand-off is a single scalar / atomic
# string so knitr's inline substitution rules apply cleanly (multi-line
# HTML is emitted as pure markdown text, not through this data chunk).
n_cohort_snv <- sum(mlst_all$dt == 1, na.rm = TRUE)
n_cohort_cna <- sum(mlst_all$dt == 4, na.rm = TRUE)
n_cohort_bnd <- sum(mlst_all$dt == 2, na.rm = TRUE)

# Drop the per-record `sample` field before serializing — DISCO's adhoc
# launcher ignores unknown keys but stripping keeps the embedded JSON minimal.
disco_mlst_prep <- lapply(mlst_all_list, function(rec) rec[names(rec) != "sample"])
disco_mlst_json <- jsonlite::toJSON(disco_mlst_prep, auto_unbox = TRUE, na = "null")

# Escape backslash + double-quote so the label is safe inside a JS string literal.
safe_cohort_label <- gsub('"', '\\"',
                          gsub("\\\\", "\\\\\\\\", as.character(cohort_label)),
                          fixed = TRUE)
```

```{r lollipop-plot-data, include = FALSE}
#| eval: !expr isTRUE(params$disco_enable) && exists("snv_long") && nrow(snv_long) > 0 && "driver_tier" %in% names(snv_long)

# Filter to Tier-1 SNVs with a resolvable gene SYMBOL that can project onto
# a coding-region protein. See earlier lollipop-plot-embed comment for the
# rationale on is_protein_projectable (CSDE1 vs NRAS at chr1:114716126 case).
tier1_snv <- dplyr::filter(snv_long, driver_tier == 1L)
pp_host   <- if (!is.null(params$proteinpaint_host)   && nzchar(params$proteinpaint_host))   params$proteinpaint_host   else "http://localhost:3456"
pp_genome <- if (!is.null(params$proteinpaint_genome) && nzchar(params$proteinpaint_genome)) params$proteinpaint_genome else "hg38"

snv_class_from_consequence <- if (exists("disco_snv_class")) {
  disco_snv_class
} else {
  function(c) "M"
}

is_protein_projectable <- function(hgvsp, consequence) {
  hgvsp_ok <- !is.na(hgvsp) & nzchar(hgvsp)
  cons_ok  <- !is.na(consequence) & nzchar(consequence) &
              grepl(paste0("missense|stop_gained|stop_lost|start_lost|",
                           "frameshift|inframe_|splice_(acceptor|donor|region)|",
                           "protein_altering|synonymous|nonsense"),
                    consequence, ignore.case = TRUE)
  hgvsp_ok | cons_ok
}

lollipop_variants <- if (nrow(tier1_snv) == 0) tibble::tibble() else {
  n_r         <- nrow(tier1_snv)
  sample_col  <- if ("sample_id" %in% names(tier1_snv)) "sample_id" else "SAMPLE"
  sample_vals <- as.character(tier1_snv[[sample_col]])
  sample_vals <- ifelse(is.na(sample_vals) | !nzchar(sample_vals), "unknown", sample_vals)
  clnsig_vals <- if ("CLNSIG" %in% names(tier1_snv)) as.character(tier1_snv$CLNSIG) else rep(NA_character_, n_r)
  onc_vals    <- if ("ONC"    %in% names(tier1_snv)) as.character(tier1_snv$ONC)    else rep(NA_character_, n_r)
  clnsig_vals[!nzchar(clnsig_vals) | clnsig_vals == "."] <- NA_character_
  onc_vals   [!nzchar(onc_vals)    | onc_vals    == "."] <- NA_character_
  pathogenicity_vals <- dplyr::coalesce(clnsig_vals, onc_vals, "")
  c_vaf <- if ("CLAIRSTO_VAF"    %in% names(tier1_snv)) suppressWarnings(as.numeric(tier1_snv$CLAIRSTO_VAF)) else rep(NA_real_, n_r)
  d_vaf <- if ("DEEPSOMATIC_VAF" %in% names(tier1_snv)) suppressWarnings(as.numeric(tier1_snv$DEEPSOMATIC_VAF)) else rep(NA_real_, n_r)
  vaf_max <- pmax(c_vaf, d_vaf, na.rm = FALSE)
  vaf_max <- ifelse(is.na(vaf_max), dplyr::coalesce(c_vaf, d_vaf), vaf_max)
  vaf_vals <- ifelse(is.na(vaf_max), "", sprintf("%.3f", vaf_max))
  consequence_vals <- if ("Consequence" %in% names(tier1_snv)) as.character(tier1_snv$Consequence) else rep("", n_r)
  hgvsp_col <- if ("HGVSp" %in% names(tier1_snv)) as.character(tier1_snv$HGVSp) else rep(NA_character_, n_r)

  tier1_snv |>
    dplyr::transmute(
      gene_symbol = dplyr::coalesce(SYMBOL, NA_character_),
      chr    = if (all(startsWith(as.character(CHROM), "chr"))) as.character(CHROM)
               else paste0("chr", CHROM),
      pos    = suppressWarnings(as.integer(POS)),
      mname  = dplyr::coalesce(HGVSp, HGVSc, ""),
      class  = snv_class_from_consequence(Consequence),
      dt     = 1L,
      ref    = REF, alt = ALT, sample = sample_vals,
      Consequence   = consequence_vals,
      Pathogenicity = pathogenicity_vals,
      VAF           = vaf_vals,
      .hgvsp_raw    = hgvsp_col
    ) |>
    dplyr::filter(!is.na(pos), !is.na(gene_symbol), nzchar(gene_symbol),
                  is_protein_projectable(.hgvsp_raw, Consequence)) |>
    dplyr::select(-.hgvsp_raw)
}

build_lollipop_variants_json <- function(variants_df) {
  info_cols <- intersect(c("Consequence", "Pathogenicity", "VAF"), names(variants_df))
  top_cols  <- setdiff(names(variants_df), c("gene_symbol", info_cols))
  is_empty_scalar <- function(v) {
    if (is.null(v)) return(TRUE)
    if (length(v) != 1) return(FALSE)
    if (is.na(v)) return(TRUE)
    if (is.character(v) && !nzchar(v)) return(TRUE)
    FALSE
  }
  vars <- lapply(seq_len(nrow(variants_df)), function(i) {
    top_part  <- as.list(variants_df[i, top_cols,  drop = FALSE])
    info_part <- as.list(variants_df[i, info_cols, drop = FALSE])
    info_part <- info_part[!vapply(info_part, is_empty_scalar, logical(1))]
    if (length(info_part) > 0) top_part$info <- info_part
    top_part
  })
  jsonlite::toJSON(vars, auto_unbox = TRUE, na = "null")
}

build_lollipop_html <- function(gene, variants_df) {
  vars_json <- build_lollipop_variants_json(variants_df)
  safe_gene <- gsub('"', '\\"', gene, fixed = TRUE)
  sprintf('<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Lollipop &mdash; %s (Tier-1 SNVs)</title>
  <style>
    body { margin: 0; padding: 12px; font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", sans-serif; background: #ffffff; }
    #lollipop_pp_holder { min-height: 500px; width: 100%%; }
  </style>
  <script src="https://proteinpaint.stjude.org/bin/proteinpaint.js" charset="utf-8"></script>
</head>
<body>
  <div id="lollipop_pp_holder"></div>
  <script>
    document.addEventListener("DOMContentLoaded", function () {
      if (typeof runproteinpaint !== "function") {
        document.getElementById("lollipop_pp_holder").innerHTML =
          "<em>ProteinPaint JS bundle failed to load. Verify network access to proteinpaint.stjude.org or host a local mirror.</em>";
        return;
      }
      runproteinpaint({
        host: "%s",
        holder: document.getElementById("lollipop_pp_holder"),
        parseurl: true, noheader: true, nobox: true,
        genome: "%s",
        gene: "%s",
        tracks: [ { type: "mds3", name: "Tier-1 SNVs %s", custom_variants: %s } ]
      });
    });
  </script>
</body>
</html>
', safe_gene, pp_host, pp_genome, safe_gene, safe_gene, vars_json)
}

# Sanitize label for filename
if (!exists("disco_sample_slug")) {
  disco_sample_slug <- function(s) {
    s <- as.character(s)
    s[is.na(s) | !nzchar(s)] <- "unknown"
    gsub("[^A-Za-z0-9._-]+", "_", s)
  }
}

lollipop_link_display <- if (nrow(lollipop_variants) == 0) tibble::tibble() else {
  dir.create("protein_paint_out", showWarnings = FALSE, recursive = TRUE)
  gene_counts <- lollipop_variants |>
    dplyr::group_by(gene_symbol) |>
    dplyr::summarise(
      n_snv       = dplyr::n(),
      top_variant = {
        vaf_num <- suppressWarnings(as.numeric(VAF))
        mname[order(-vaf_num, na.last = TRUE)][1]
      },
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(n_snv), gene_symbol)

  lollipop_files <- character(nrow(gene_counts))
  for (i in seq_len(nrow(gene_counts))) {
    g   <- gene_counts$gene_symbol[i]
    vg  <- dplyr::filter(lollipop_variants, gene_symbol == g)
    out <- file.path("protein_paint_out",
                     sprintf("lollipop_%s.html", disco_sample_slug(g)))
    writeLines(build_lollipop_html(g, vg), out)
    lollipop_files[i] <- out
  }
  tibble::tibble(
    Gene         = sprintf('<a href="%s" target="_blank" rel="noopener">%s</a>',
                           lollipop_files, gene_counts$gene_symbol),
    `# Tier-1`   = gene_counts$n_snv,
    `Top variant` = ifelse(is.na(gene_counts$top_variant) | !nzchar(gene_counts$top_variant),
                          "&mdash;", gene_counts$top_variant),
    Open         = sprintf('<a href="%s" target="_blank" rel="noopener">&#128279;</a>',
                          lollipop_files)
  )
}
```





# Overview

## Row {height="120px"}

```{r vb-samples}
#| content: valuebox
#| title: "Samples"
list(icon = "people", color = "primary",
     value = vb_count("stats_df", function(df) rep(TRUE, nrow(df))))
```

```{r vb-t1-sv}
#| content: valuebox
#| title: "Tier-1 SVs"
list(icon = "diagram-3", color = "success",
     value = vb_count("pass2_clean", function(df) df$driver_tier == 1L))
```

```{r vb-t1-cna}
#| content: valuebox
#| title: "Tier-1 CNAs"
list(icon = "bar-chart", color = "warning",
     value = vb_count("cna_pass2_clean", function(df) df$driver_tier == 1L))
```

```{r vb-t1-snv}
#| content: valuebox
#| title: "Tier-1 SNVs"
list(icon = "bullseye", color = "info",
     value = vb_count("snv_long", function(df) df$driver_tier == 1L))
```

```{r vb-total-pass2}
#| content: valuebox
#| title: "Total pass2"
list(icon = "clipboard-data", color = "secondary",
     value = vb_count("pass2_clean", function(df) rep(TRUE, nrow(df))) +
             vb_count("cna_pass2_clean", function(df) rep(TRUE, nrow(df))) +
             vb_count("snv_long", function(df) rep(TRUE, nrow(df))))
```

## Row {.fill}

### Cohort DISCO plot {title="Cohort DISCO plot"}

::: {#disco_pp_cohort style="min-height:700px; width:100%;"}
:::

<script src="https://proteinpaint.stjude.org/bin/proteinpaint.js" charset="utf-8"></script>
<script>
runproteinpaint({
  host: "`r pp_host`",
  holder: document.getElementById("disco_pp_cohort"),
  noheader: true,
  nobox: true,
  genome: "`r pp_genome`",
  disco: {
    mlst: `r disco_mlst_json`,
    settings: { Disco: { centerText: "`r safe_cohort_label`" } }
  }
});
</script>

::: {style="font-size:0.85em; color:#666; margin-top:6px;"}
Cohort DISCO plot &mdash; `r n_cohort_snv` SNV / `r n_cohort_cna` CNA / `r n_cohort_bnd` BND across `r n_cohort_samples` sample(s). Backend: `` `r pp_host` ``.
:::



```{r disco-plot-sample-statement, results = "asis", echo = FALSE}
#| eval: !expr isTRUE(params$disco_enable) && exists("sample_ids") && length(sample_ids) > 0

# ── Per-sample DISCO HTMLs are written to protein_paint_out/ by the data
# chunk above. The report body no longer renders per-sample link cards
# (they became cluttered on larger cohorts); this asis chunk emits one
# short statement pointing reviewers at the output directory instead.
# Kept as an asis chunk (not inline r) because the sample count is one
# among several substitutions in a formatted <p> block — a single inline
# `r sample_count` in prose reads less clearly than a scoped cat().
cat(sprintf(
'<p style="margin:14px 0 4px 0; font-size:0.92em; color:#444;">
  Per-sample DISCO HTMLs (%d file(s)) are written to
  <code style="background:#f2f2f2; padding:1px 6px; border-radius:3px;">%s/</code>
  under <code style="background:#f2f2f2; padding:1px 6px; border-radius:3px;">results/sv_report/</code>.
  Open the individual files directly to inspect a single sample.
</p>
', length(sample_ids), "protein_paint_out"))
```


## Row {.fill}

### SV Protein Paint {title="SV Protein Paint (gene fusion)"}

```{r sv-proteinpaint-build}
#| echo: false
#| warning: false

# ── ProteinPaint svfusion export ────────────────────────────────────────────
# Renders Tier 1 + Tier 2 SVs as a two-breakpoint fusion table for the
# ProteinPaint gene-fusion app (https://proteinpaint.stjude.org/?appcard=genefusion,
# or the FusionEditor at /FusionEditor/). Format spec:
# https://docs.stjude.cloud/visualization-community/proteinpaint/sv-and-fusion-transcript
#
# THREE THINGS THE SPEC REQUIRES THAT THE PASS2 TABLE DOES NOT ALREADY SATISFY:
#   1. positions are 0-BASED   -> POS - 1 (VCF POS is 1-based)
#   2. chromosomes need "chr"  -> raw CHROM is bare ("21")
#   3. one row per EVENT       -> an interchromosomal BND is TWO VCF records
#                                 (both breakends); emitting both duplicates it
#
# 5'/3' ASSIGNMENT comes from the ALT bracket orientation, NOT from POS order.
# VCF 4.2 breakend forms:  t[p[ and t]p]  -> THIS record is 5'
#                          ]p]t and [p[t  -> the PARTNER is 5'
# i.e. an ALT beginning with the ref base means this side is 5'. Both mates of a
# pair independently agree on the assignment, so dedup may keep either one.
#
# INS is excluded: END == POS gives no second breakpoint, and ProteinPaint's
# model is two joined breakpoints. (The spec's "inserted segment" is a different
# concept — sequence between two breakpoints in an RNA-seq fusion.)
pp_sv_fusion      <- NULL
pp_sv_note        <- NULL
pp_sv_input_json  <- "null"   # JS-safe string of the whole svfusion file

.need <- c("SAMPLE", "CHROM", "POS", "END", "ALT", "SVTYPE", "driver_tier")
if (exists("pass2_clean") && is.data.frame(pass2_clean) &&
    nrow(pass2_clean) > 0 && all(.need %in% names(pass2_clean))) {

  .pfx <- function(x) ifelse(is.na(x) | !nzchar(x) | startsWith(x, "chr"),
                             x, paste0("chr", x))
  # gene / isoform / strand per side come from the EnsDb join above. They are
  # absent when params$ensembl_db is unset — fall back to blanks, which the
  # spec permits (renders an intergenic box rather than a protein).
  .col <- function(nm) if (nm %in% names(pass2_clean)) as.character(pass2_clean[[nm]]) else NA_character_

  .d <- tibble::tibble(
    SAMPLE = as.character(pass2_clean$SAMPLE),
    CHROM  = as.character(pass2_clean$CHROM),
    POS    = suppressWarnings(as.numeric(pass2_clean$POS)),
    END    = suppressWarnings(as.numeric(pass2_clean$END)),
    ALT    = as.character(pass2_clean$ALT),
    SVTYPE = as.character(pass2_clean$SVTYPE),
    tier   = suppressWarnings(as.integer(pass2_clean$driver_tier)),
    g_l    = .col("gene_name_left"),            g_r = .col("gene_name_right"),
    t_l    = .col("canonical_transcript_left"), t_r = .col("canonical_transcript_right"),
    s_l    = .col("gene_strand_left"),          s_r = .col("gene_strand_right"),
    DV     = .col("DV")
  ) |>
    dplyr::filter(!is.na(tier), tier %in% c(1L, 2L)) |>
    dplyr::filter(SVTYPE != "INS")

  if (nrow(.d) > 0) {
    .m     <- stringr::str_match(.d$ALT, "[\\[\\]]([^:\\[\\]]+):([0-9]+)[\\[\\]]")
    .isbnd <- .d$SVTYPE == "BND" | grepl("[\\[\\]]", .d$ALT)
    .ok    <- !is.na(.m[, 2])
    .five  <- grepl("^[ACGTNacgtn]", .d$ALT)      # this record is the 5' side

    .d <- .d |>
      dplyr::mutate(
        .bnd = .isbnd & .ok,
        chr_a = .pfx(ifelse(.bnd & !.five, .m[, 2], CHROM)),
        pos_a = ifelse(.bnd & !.five, suppressWarnings(as.numeric(.m[, 3])), POS),
        gen_a = ifelse(.bnd & !.five, g_r, g_l),
        iso_a = ifelse(.bnd & !.five, t_r, t_l),
        str_a = ifelse(.bnd & !.five, s_r, s_l),
        chr_b = .pfx(ifelse(.bnd, ifelse(.five, .m[, 2], CHROM), CHROM)),
        pos_b = ifelse(.bnd, ifelse(.five, suppressWarnings(as.numeric(.m[, 3])), POS), END),
        gen_b = ifelse(.bnd & !.five, g_l, g_r),
        iso_b = ifelse(.bnd & !.five, t_l, t_r),
        str_b = ifelse(.bnd & !.five, s_l, s_r)
      ) |>
      # drop non-BND records that have no usable second breakpoint
      dplyr::filter(!is.na(pos_a), !is.na(pos_b), pos_a != pos_b | .bnd) |>
      # one row per event: collapse the two breakends of a BND pair
      dplyr::mutate(.key = paste(SAMPLE,
                                 pmin(paste0(chr_a, ":", pos_a), paste0(chr_b, ":", pos_b)),
                                 pmax(paste0(chr_a, ":", pos_a), paste0(chr_b, ":", pos_b)))) |>
      dplyr::distinct(.key, .keep_all = TRUE)

    pp_sv_fusion <- .d |>
      dplyr::transmute(
        gene_a = dplyr::coalesce(gen_a, ""), isoform_a = dplyr::coalesce(iso_a, ""),
        chr_a, position_a = as.integer(pos_a - 1), strand_a = dplyr::coalesce(str_a, "+"),
        gene_b = dplyr::coalesce(gen_b, ""), isoform_b = dplyr::coalesce(iso_b, ""),
        chr_b, position_b = as.integer(pos_b - 1), strand_b = dplyr::coalesce(str_b, "+"),
        Sample = SAMPLE, Sampletype = "diagnosis",
        SVTYPE, driver_tier = tier, Discordant_readcount = dplyr::coalesce(DV, "")
      ) |>
      dplyr::arrange(Sample, chr_a, position_a)

    dir.create("protein_paint_out", showWarnings = FALSE, recursive = TRUE)
    # Fixed descriptive name, matching the disco_plot_*/lollipop_* convention
    # used elsewhere in protein_paint_out/ (no PREFIX variable exists here).
    .out <- file.path("protein_paint_out", "sv_tier1_2.svfusion.txt")
    readr::write_tsv(pp_sv_fusion, .out, na = "")
    # fusioneditor expects `input` = the ENTIRE file as one string, "\n" line
    # breaks, tab-separated, header first. toJSON(auto_unbox) emits a correctly
    # escaped JS string literal.
    pp_sv_input_json <- as.character(jsonlite::toJSON(
      paste(readLines(.out, warn = FALSE), collapse = "\n"), auto_unbox = TRUE))
    pp_sv_note <- sprintf("%d event(s) across %d sample(s) &rarr; <code>%s</code>",
                          nrow(pp_sv_fusion),
                          dplyr::n_distinct(pp_sv_fusion$Sample), .out)
  } else {
    pp_sv_note <- "No Tier 1/2 SVs with two breakpoints."
  }
} else {
  pp_sv_note <- "pass2_clean unavailable or missing required columns."
}
```

```{r sv-proteinpaint-table}
#| echo: false
if (!is.null(pp_sv_fusion) && nrow(pp_sv_fusion) > 0) {
  DT::datatable(
    pp_sv_fusion,
    rownames   = FALSE,
    extensions = c("Buttons", "Scroller"),
    options = list(
      dom = "Bfrtip", buttons = list("copy", list(extend = "csv", text = "Download svfusion")),
      scrollY = 360, scroller = TRUE, scrollX = TRUE, pageLength = 25
    ),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-size:0.9em;",
      htmltools::HTML(paste0(
        "ProteinPaint <b>svfusion</b> input &mdash; Tier 1 + Tier 2, one row per event. ",
        "Positions are <b>0-based</b> per the format spec. ",
        "Load at <code>https://proteinpaint.stjude.org/FusionEditor/</code> ",
        "or the <code>?appcard=genefusion</code> card."))))
} else {
  htmltools::HTML(paste0("<p style='color:#666;'>", pp_sv_note, "</p>"))
}
```

::: {#sv_fusion_pp style="min-height:520px; width:100%; margin-top:10px;"}
:::

<script src="https://proteinpaint.stjude.org/bin/proteinpaint.js" charset="utf-8"></script>
<script>
// Fusion Editor embed. Key names are case-sensitive; `input` (NOT `text`) is the
// documented parameter and takes the entire file content as one string.
// Ref: Embedding.md "Fusion editor" / fusioneditor:{}
//      live example https://proteinpaint.stjude.org/examples/fusioneditor.html
(function () {
  var ppInput = `r pp_sv_input_json`;
  if (!ppInput) return;                       // no Tier 1/2 events -> render nothing
  runproteinpaint({
    host: "`r pp_host`",
    holder: document.getElementById("sv_fusion_pp"),
    noheader: true,
    nobox: true,
    genome: "`r pp_genome`",
    fusioneditor: {
      input: ppInput,
      dataname: "`r safe_cohort_label` Tier 1/2 SVs"
    }
  });
})();
</script>

::: {style="font-size:0.85em; color:#666; margin-top:6px;"}
SV Protein Paint &mdash; `r if (!is.null(pp_sv_note)) htmltools::HTML(pp_sv_note) else ""`.
Genome build `r pp_genome`. BND mate pairs are collapsed to one event; INS is excluded
(no second breakpoint).
:::

## Row {.fill}

### Cohort recurrent SVs {title="Cohort recurrent SVs (Jasmine merge)"}

```{r cohort-merge-load}
#| echo: false
#| warning: false

# ── COHORT_MERGE_SV results ─────────────────────────────────────────────────
# Populated only on PASS 2 of the two-pass render (see the header of
# modules/local/sv_report_index/main.nf). On pass 1 params$cohort_recurrent_tsv
# is "" and this section renders an explanatory card: pass 1's report is
# superseded by pass 2 whenever the merge runs, so an empty card there is
# expected, not a failure.
#
# The recurrence table is LONG -- one row per (cluster x sample). Two counts
# carry the interpretation (see modules/local/jasmine_cluster_map/main.nf):
#   n_tiered = samples where this SV was a Tier 1/2/3 call
#   n_any    = samples with ANY evidence in the cluster (Jasmine INFO/SUPP)
# n_tiered=1, n_any>=2 -> RESCUE candidate: called confidently in one sample and
#                         filtered out elsewhere despite supporting evidence.
# n_tiered=1, n_any=1  -> no corroboration; private call or false positive.
jm_recur    <- NULL
jm_note     <- NULL
jm_summary  <- NULL
jm_hdr_note <- NULL

# bgzf is gzip-readable, so the header can be pulled without an htslib dependency.
# Reads line-by-line and stops at the first non-"##" line rather than slurping a
# potentially large VCF body.
.read_vcf_header <- function(path, cap = 20000L) {
  con <- gzfile(path, "rt")
  on.exit(try(close(con), silent = TRUE), add = TRUE)
  out <- character(0)
  repeat {
    l <- readLines(con, n = 1L, warn = FALSE)
    if (length(l) == 0L || !startsWith(l, "##")) break
    out <- c(out, l)
    if (length(out) >= cap) break
  }
  out
}

.crt <- params$cohort_recurrent_tsv
if (is.character(.crt) && length(.crt) == 1L && nzchar(.crt) && file.exists(.crt)) {
  .tbl <- try(suppressWarnings(readr::read_tsv(
    .crt, show_col_types = FALSE, progress = FALSE)), silent = TRUE)
  if (inherits(.tbl, "try-error")) {
    jm_note <- sprintf("Could not read <code>%s</code>.", basename(.crt))
  } else if (nrow(.tbl) == 0) {
    jm_note <- paste0(
      "COHORT_MERGE_SV ran, but no cluster reached the recurrence threshold ",
      "(<code>cohort_sv_min_recurrence</code>). Expected on a two-sample test set.")
  } else {
    jm_recur <- .tbl
  }
} else {
  jm_note <- paste0(
    "Cohort SV merge results are not part of this render. Either this is ",
    "<b>pass 1</b> of the two-pass render &mdash; it produces the tiered table ",
    "that COHORT_MERGE_SV consumes, and pass 2 re-renders this page with the ",
    "merge results &mdash; or <code>run_cohort_merge_sv</code> is false.")
}

if (!is.null(jm_recur)) {
  .cols <- names(jm_recur)
  if (all(c("cluster_id", "n_tiered", "n_any") %in% .cols)) {
    jm_summary <- jm_recur |>
      dplyr::distinct(cluster_id, .keep_all = TRUE) |>
      dplyr::mutate(.nt = suppressWarnings(as.integer(n_tiered)),
                    .na = suppressWarnings(as.integer(n_any))) |>
      dplyr::summarise(
        clusters              = dplyr::n(),
        `n_any >= 2`          = sum(.na >= 2, na.rm = TRUE),
        `n_tiered >= 2`       = sum(.nt >= 2, na.rm = TRUE),
        `rescue candidates`   = sum(.nt == 1 & .na >= 2, na.rm = TRUE))
    # Most-corroborated clusters first -- the rows a reviewer looks at.
    .ord <- intersect(c("cluster_id", "sample"), .cols)
    jm_recur <- jm_recur |>
      dplyr::arrange(dplyr::desc(suppressWarnings(as.integer(n_any))),
                     dplyr::across(dplyr::all_of(.ord)))
  }
}

.cmv <- params$cohort_merged_vcf
if (is.character(.cmv) && length(.cmv) == 1L && nzchar(.cmv) && file.exists(.cmv)) {
  .hdr <- try(.read_vcf_header(.cmv), silent = TRUE)
  if (!inherits(.hdr, "try-error") && length(.hdr) > 0) {
    # Jasmine copies the header of its FIRST input, which here is a fully
    # VEP/AnnotSV-annotated per-sample VCF, so the upstream INFO definitions are
    # expected to survive the merge. Stated as a CHECK, not an assumption: the
    # card below reports which tags the merged file actually declares. Note that
    # a declared INFO tag is not the same as a populated one -- Jasmine emits one
    # representative record per cluster, so member-level annotations are taken
    # from the per-sample record table by JASMINE_CLUSTER_MAP, not from here.
    # NOT "CSQ": SPLIT_VEP_TO_INFO expands it into individual INFO fields
    # upstream, so CSQ is already absent from Jasmine's INPUT.
    .want <- c("Gene_name", "gnomAD_SV_AF", "ClinVar_SV_CLNSIG", "CALLER",
               "AnnotSV_ID", "ACMG_class", "AnnotSV_ranking_score",
               "SUPP_VEC", "SUPP", "IDLIST", "STARTVARIANCE")
    .have <- vapply(.want, function(k)
      any(grepl(sprintf("^##INFO=<ID=%s,", k), .hdr, perl = TRUE)), logical(1))
    jm_hdr_note <- sprintf(
      paste0("Merged VCF <code>%s</code> &mdash; %d header meta-lines. ",
             "INFO declared: <b>%s</b>. Absent: %s."),
      basename(.cmv), length(.hdr),
      if (any(.have))  paste(names(.have)[.have],  collapse = ", ") else "none",
      if (any(!.have)) paste(names(.have)[!.have], collapse = ", ") else "none")
  }
}
```

```{r cohort-merge-summary}
#| echo: false
if (!is.null(jm_summary)) {
  knitr::kable(jm_summary, format = "html",
               table.attr = 'class="table table-sm" style="width:auto;"')
}
```

```{r cohort-merge-table}
#| echo: false
#| column: screen-inset
if (!is.null(jm_recur) && nrow(jm_recur) > 0) {
  DT::datatable(
    jm_recur,
    escape     = FALSE,
    filter     = "top",
    rownames   = FALSE,
    width      = "100%",
    extensions = c("Buttons", "Scroller"),
    options = list(
      dom     = "Bfrtip",
      buttons = list("copy", list(extend = "csv", text = "Download recurrence TSV")),
      scrollY = 420, scroller = TRUE, scrollX = TRUE, pageLength = 25
    ),
    caption = htmltools::tags$caption(
      style = "caption-side: top; text-align: left; font-size:0.9em;",
      htmltools::HTML(paste0(
        "Jasmine cohort merge &mdash; one row per (cluster &times; sample). ",
        "<code>record_class</code> is TIERED when the row is in this report's ",
        "tiered table, and RESCUED when it is a sub-threshold call recovered ",
        "from the pre-filter VEP VCF. <code>n_tiered</code> against ",
        "<code>n_any</code> separates over-filtering from lack of ",
        "corroboration."))))
} else {
  htmltools::HTML(paste0("<p style='color:#666; margin:10px 0;'>", jm_note, "</p>"))
}
```

::: {style="font-size:0.85em; color:#666; margin-top:6px;"}
`r if (!is.null(jm_hdr_note)) htmltools::HTML(jm_hdr_note) else "Merged VCF not staged in this render."`
:::

## Row {height="240px"}

### Per-sample AnnotSV summary {title="Per-sample AnnotSV summary"}

```{r sample-stats-table}
#| column: screen-inset
DT::datatable(
  stats_df,
  escape     = FALSE,
  filter     = "top",
  rownames   = FALSE,
  width      = "100%",
  extensions = "Buttons",
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 25,
    scrollX    = TRUE,
    autoWidth  = TRUE
  )
)
```


# SV

## Row {height="45%" .tabset}

### SVTYPE {title="SVTYPE distribution"}

```{r plot-svtype}
#| eval: !expr nrow(pass2_long) > 0

p1 <- pass2_clean |>
  count(SAMPLE, SVTYPE, name = "n") |>
  ggplot(aes(x = SAMPLE, y = n, fill = SVTYPE,
             text = paste0("<b>Sample:</b> ", SAMPLE,
                           "<br><b>SVTYPE:</b> ", SVTYPE,
                           "<br><b>Variants (pass2):</b> ", n))) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "viridis", end = 0.9) +
  labs(title    = "Pass2 variants by SVTYPE",
       x = NULL, y = "Variant count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        axis.text.x     = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplotly(p1, tooltip = "text") |> config(displayModeBar = FALSE)
```

### ACMG {title="ACMG class + AnnotSV ranking score"}

```{r plot-acmg}
#| eval: !expr nrow(pass2_long) > 0

p2a <- pass2_clean |>
  count(SAMPLE, ACMG_class_clean, name = "n") |>
  ggplot(aes(x = SAMPLE, y = n, fill = ACMG_class_clean,
             text = paste0("<b>Sample:</b> ", SAMPLE,
                           "<br><b>ACMG class:</b> ", ACMG_class_clean,
                           "<br><b>Variants:</b> ", n))) +
  geom_col(position = "stack") +
  scale_fill_manual(values = acmg_palette, name = "ACMG class",
                    drop = FALSE, na.value = "grey70") +
  labs(title    = "ACMG class distribution",
       subtitle = "counts of variants per sample, colored by ACMG ranking (NA = grey)",
       x = NULL, y = "Variant count") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# Plot the AnnotSV ranking score as jittered points (no violin / density).
# geom_violin crashes plotly's to_basic.GeomViolin when the layer's internal data frame
# has too few rows ("argument 1 is not a vector" from order(data[["y"]])).
# geom_jitter has no density step and works for any N ≥ 1.
p2b_data <- pass2_clean |>
  filter(!is.na(ranking_score), ranking_score > 0)

if (nrow(p2b_data) > 0) {
  p2b <- p2b_data |>
    ggplot(aes(x = SAMPLE, y = ranking_score, color = SAMPLE,
               text = paste0("<b>Sample:</b> ", SAMPLE,
                             "<br><b>ID:</b> ", ID,
                             "<br><b>SVTYPE:</b> ", SVTYPE,
                             "<br><b>AnnotSV ranking score:</b> ", ranking_score))) +
    geom_jitter(width = 0.15, height = 0, alpha = 0.8, size = 3) +
    scale_color_viridis_d(option = "mako", end = 0.85, guide = "none") +
    labs(title    = "AnnotSV ranking score",
         subtitle = "Per-variant AnnotSV ranking score (jittered) — only records with score > 0 shown",
         x = NULL, y = "Score") +
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

  subplot(ggplotly(p2a, tooltip = "text") |> config(displayModeBar = FALSE),
          ggplotly(p2b, tooltip = "text") |> config(displayModeBar = FALSE),
          nrows = 1, titleY = TRUE, margin = 0.06)
} else {
  ggplotly(p2a, tooltip = "text") |> config(displayModeBar = FALSE)
}
```

### ST17 {title="ST17 recurrence hits by SVTYPE"}

```{r plot-st17}
#| eval: !expr nrow(pass2_long) > 0

# Map each SVTYPE. the matching name_ST17_<SVTYPE> column
st17_cols <- c(BND = "name_ST17_BND",
               DEL = "name_ST17_DEL",
               DUP = "name_ST17_DUP",
               INS = "name_ST17_INS",
               INV = "name_ST17_INV")

# rlang-patterns: parameterize over the (SVTYPE, ST17 column) pair using
# tidyselect's .data[[ ]] to look up the column by string name per group.
# Uses pass2_clean (not pass2_long) so ST17 SVTYPE-mismatch blanking and
# BND partner-chromosome validation are already applied.
plot3_data <- imap_dfr(st17_cols, function(col_name, svtype) {
  if (!col_name %in% names(pass2_clean)) return(tibble())
  pass2_clean |>
    filter(SVTYPE == svtype) |>
    mutate(has_st17_hit = is_meaningful(.data[[col_name]])) |>
    count(SAMPLE, has_st17_hit, name = "n") |>
    mutate(SVTYPE = svtype, .before = 1)
})

p3 <- ggplot(plot3_data,
             aes(x = SAMPLE, y = n, fill = has_st17_hit,
                 text = paste0("<b>Sample:</b> ", SAMPLE,
                               "<br><b>SVTYPE:</b> ", SVTYPE,
                               "<br><b>", ifelse(has_st17_hit, "ST17 Hit", "ST17 Miss"), "</b>",
                               "<br><b>Variants:</b> ", n))) +
  geom_col(position = "stack") +
  facet_wrap(~ SVTYPE, nrow = 1, scales = "free_y") +
  scale_fill_viridis_d(option = "viridis", begin = 0.2, end = 0.85,
                       name = "ST17 hit?",
                       labels = c(`TRUE` = "Hit", `FALSE` = "Miss")) +
  labs(title    = "ST17 recurrence-BED hits by SVTYPE",
       subtitle = "Per-sample variants overlapping the ST17 T-ALL recurrence BED, by SV type",
       x = NULL, y = "Variant count") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplotly(p3, tooltip = "text") |> config(displayModeBar = FALSE)
```

### VAF {title="VAF / AF metric distributions"}

```{r plot-vaf-dist}
#| eval: !expr nrow(pass2_long) > 0


# somatic-signal. gnomAD_SV_AF and TUMOUR_AF are comma-joined multi-overlap tuples; take the max as the population-frequency / tumor-signal
parse_max_af <- function(x) {
  if (is.na(x) || !nzchar(x)) return(NA_real_)
  vals <- suppressWarnings(as.numeric(str_split(x, ",")[[1]]))
  vals <- vals[!is.na(vals)]
  if (length(vals) == 0) NA_real_ else max(vals)
}

vaf_metrics <- pass2_clean |>
  mutate(
    hVAF_HP1      = suppressWarnings(as.numeric(str_split_fixed(hVAF, ",", 3)[, 1])),
    hVAF_HP2      = suppressWarnings(as.numeric(str_split_fixed(hVAF, ",", 3)[, 2])),
    VAF_num       = suppressWarnings(as.numeric(VAF)),
    gnomAD_AF_max = vapply(gnomAD_SV_AF, parse_max_af, numeric(1), USE.NAMES = FALSE),
    TUMOUR_AF_max = vapply(TUMOUR_AF,    parse_max_af, numeric(1), USE.NAMES = FALSE)
  ) |>
  select(SAMPLE, ID, SVTYPE, driver_tier,
         `Severus VAF`            = VAF_num,
         `Severus hVAF (HP1)`     = hVAF_HP1,
         `Severus hVAF (HP2)`     = hVAF_HP2,
         `SAVANA TUMOUR_AF (max)` = TUMOUR_AF_max,
         `gnomAD_SV_AF (max)`     = gnomAD_AF_max) |>
  pivot_longer(cols      = -c(SAMPLE, ID, SVTYPE, driver_tier),
               names_to  = "metric",
               values_to = "value") |>
  filter(!is.na(value))

p_vaf <- vaf_metrics |>
  ggplot(aes(x = metric, y = value,
             color = as.factor(driver_tier),
             text  = paste0("<b>Sample:</b> ",  SAMPLE,
                            "<br><b>ID:</b> ",    ID,
                            "<br><b>SVTYPE:</b> ", SVTYPE,
                            "<br><b>Driver tier:</b> T", driver_tier,
                            "<br><b>", metric, ":</b> ", round(value, 4)))) +
  geom_boxplot(outlier.shape = NA, alpha = 0.15,
               color = "grey40", fill = "grey85", show.legend = FALSE) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.75, size = 2) +
  scale_color_manual(values = c(`1` = "#2CA02C", `2` = "#D68910", `3` = "#6c757d"),
                     name   = "Driver tier",
                     labels = c(`1` = "T1 (ST17)",
                                `2` = "T2 (Candidate)",
                                `3` = "T3 (Other)")) +
  labs(title    = "VAF / AF metric distributions",
       subtitle = "Boxplots summarize distribution; points are individual pass2 SVs colored by driver_tier",
       x = NULL, y = "Allele frequency") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_vaf, tooltip = "text") |> config(displayModeBar = FALSE)
```

### Read depth {title="Supporting read-depth distributions"}

```{r plot-read-support}
#| eval: !expr nrow(pass2_long) > 0

# Three read-support metrics — Severus reference reads (DR), Severus variant reads (DV), and SAVANA TUMOUR_READ_SUPPORT. Higher DV / TUMOUR_READ_SUPPORT. higher confidence in the somatic call. 
read_metrics <- pass2_clean |>
  mutate(across(c(DR, DV, TUMOUR_READ_SUPPORT),
                ~ suppressWarnings(as.numeric(.x)))) |>
  select(SAMPLE, ID, SVTYPE, driver_tier,
         `Severus DR (ref reads)`      = DR,
         `Severus DV (variant reads)`  = DV,
         `SAVANA TUMOUR_READ_SUPPORT`  = TUMOUR_READ_SUPPORT) |>
  pivot_longer(cols      = -c(SAMPLE, ID, SVTYPE, driver_tier),
               names_to  = "metric",
               values_to = "reads") |>
  filter(!is.na(reads), reads > 0)

p_reads <- read_metrics |>
  ggplot(aes(x = metric, y = reads, color = SVTYPE,
             text = paste0("<b>Sample:</b> ",  SAMPLE,
                           "<br><b>ID:</b> ",    ID,
                           "<br><b>SVTYPE:</b> ", SVTYPE,
                           "<br><b>Driver tier:</b> T", driver_tier,
                           "<br><b>", metric, ":</b> ", reads, " reads"))) +
  geom_boxplot(outlier.shape = NA, alpha = 0.15,
               color = "grey40", fill = "grey85", show.legend = FALSE) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.75, size = 2) +
  scale_color_manual(values = svtype_colors, name = "SVTYPE") +
  scale_y_log10() +
  labs(title    = "Supporting read-depth distributions",
       subtitle = "Boxplots per read-support metric; points are pass2 SVs colored by SVTYPE (log10 y-axis)",
       x = NULL, y = "Read count (log10)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

ggplotly(p_reads, tooltip = "text") |> config(displayModeBar = FALSE)
```

### Driver tier {title="Driver-tier prioritized variants per sample"}

```{r plot-driver-tier}
#| eval: !expr nrow(pass2_long) > 0

# One bar per sample, stacked by driver_tier.
tier_counts <- pass2_clean |>
  count(SAMPLE, driver_tier, name = "n", .drop = FALSE) |>
  mutate(driver_tier = factor(paste0("T", driver_tier), levels = c("T1", "T2", "T3")))

p_tier <- tier_counts |>
  ggplot(aes(x = SAMPLE, y = n, fill = driver_tier,
             text = paste0("<b>Sample:</b> ",  SAMPLE,
                           "<br><b>Driver tier:</b> ", driver_tier,
                           "<br><b>Variants:</b> ",    n))) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(T1 = "#2CA02C", T2 = "#D68910", T3 = "#6c757d"),
                    name = "Driver tier",
                    labels = c(T1 = "T1 — ST17 hit",
                               T2 = "T2 — Candidate gene",
                               T3 = "T3 — Other pass2 evidence")) +
  labs(title    = "Driver-tier prioritized pass2 variants per sample",
       subtitle = "T1 = SVTYPE-matched ST17 recurrence (empirical T-ALL truth); T2 = candidate-gene hit; T3 = other pass2 evidence",
       x = NULL, y = "Variant count") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggplotly(p_tier, tooltip = "text") |> config(displayModeBar = FALSE)
```





## Row {height="55%"}

### SV pass2 table {title="SV pass2 variant table" fill="true"}

```{r pass2-table}
#| eval: !expr nrow(pass2_long) > 0
#| column: screen-inset

# Display columns for the DT: full schema minus RE_gene
# driver prioritization.
display_cols <- c(
  "SAMPLE", "sample_id", "driver_tier", "SVTYPE", "DETAILED_TYPE", "ALT",
  "breakpoint", "CytoBand_display", "has_mate", "SVLEN", "CALLER",
  "Gene_name_display", "Candidate_genes", "ST17_hit",
  "AnnotSV_ranking_score", "ACMG_class",
  "VAF", "hVAF", "DR", "DV", "TUMOUR_READ_SUPPORT", "TUMOUR_AF",
  "ClinVar_SV_CLNSIG"
)
# sample_id (friendly cohort label) is joined in from sample_metadata via
# merge_sample_metadata() earlier — silently absent when no metadata CSV.
display_cols <- intersect(display_cols, names(pass2_clean))

# Sort: driver_tier ascending (T1 first), tier_score descending. higher-recurrence / more-candidates / higher-AnnotSV-rank first 


pass2_display <- pass2_clean |>
  arrange(driver_tier, desc(tier_score), SVTYPE, SAMPLE) |>
  select(all_of(display_cols))

# svtype_colors palette is defined in the load-pass2 chunk
svtype_col_js <- which(names(pass2_display) == "SVTYPE") - 1L

# Build a JS render callback that wraps each SVTYPE cell value in a
# pill-shaped span with an INLINE style attribute.
# overrides cyborg theme's dark-mode cell defaults AND the broader white-bg
# `!important` rules in the dt-readability chunk above
svtype_render_js <- DT::JS(sprintf(
  "function(data, type, row) {
     if (type !== 'display' || data == null || data === '') return data;
     var colors = %s;
     var c = colors[data] || '#6c757d';
     return '<span style=\"background-color:' + c + ';color:#ffffff;' +
            'font-weight:bold;padding:2px 8px;border-radius:4px;' +
            'display:inline-block;min-width:36px;text-align:center;\">' +
            data + '</span>';
   }",
  jsonlite::toJSON(as.list(svtype_colors), auto_unbox = TRUE)
))

# driver_tier badge palette  
tier_colors <- c(`1` = "#2CA02C", `2` = "#D68910", `3` = "#6c757d")
tier_col_js <- which(names(pass2_display) == "driver_tier") - 1L
tier_render_js <- DT::JS(sprintf(
  "function(data, type, row) {
     if (type !== 'display' || data == null || data === '') return data;
     var colors = %s;
     var c = colors[String(data)] || '#6c757d';
     return '<span style=\"background-color:' + c + ';color:#ffffff;' +
            'font-weight:bold;padding:2px 10px;border-radius:10px;' +
            'display:inline-block;min-width:30px;text-align:center;\">' +
            'T' + data + '</span>';
   }",
  jsonlite::toJSON(as.list(tier_colors), auto_unbox = TRUE)
))

DT::datatable(
  pass2_display,
  escape     = FALSE,
  filter     = "top",
  rownames   = FALSE,
  width      = "100%",
  extensions = c("Buttons", "RowGroup"),
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 15,
    scrollX    = TRUE,
    autoWidth  = TRUE,
    # Group rows by driver_tier — one T1/T2/T3 header bar.
    rowGroup   = list(dataSrc = tier_col_js),
    # Initial sort: driver_tier ASC keeps rowGroup buckets contiguous; secondary sort of AnnotSV_ranking_score DESC within each tier 
    order      = list(list(tier_col_js, "asc")),
    columnDefs = list(
      list(targets = svtype_col_js, render = svtype_render_js),
      list(targets = tier_col_js,   render = tier_render_js)
    )
  )
)
```

# CNA

## Row {height="120px"}

```{r vb-cna-t1}
#| content: valuebox
#| title: "Tier-1 CNAs"
list(icon = "bullseye", color = "success",
     value = vb_count("cna_pass2_clean", function(df) df$driver_tier == 1L))
```

```{r vb-cna-t2}
#| content: valuebox
#| title: "Tier-2 CNAs"
list(icon = "geo-alt", color = "primary",
     value = vb_count("cna_pass2_clean", function(df) df$driver_tier == 2L))
```

```{r vb-cna-t3}
#| content: valuebox
#| title: "Tier-3 CNAs"
list(icon = "paperclip", color = "secondary",
     value = vb_count("cna_pass2_clean", function(df) df$driver_tier == 3L))
```

```{r vb-cna-unhap}
#| content: valuebox
#| title: "Unhaplotyped"
list(icon = "exclamation-triangle", color = "warning",
     value = vb_count("cna_pass2_clean", function(df) {
       (suppressWarnings(as.integer(df$CN1)) %in% 0L) & (suppressWarnings(as.integer(df$CN2)) %in% 0L)
     }))
```

## Row {.fill}

### CNA pass2 table {title="CNA pass2 variant table" fill="true"}



```{r cna-pass2-status}
if (nrow(cna_pass2_long) == 0) {
  cat("**No pass2 CNV variants found.** Either FILTER_CNA produced no records ",
      "matching the prioritization gate for any sample, or the input was not ",
      "staged. Table below is suppressed.", sep = "")
}
```



```{r cna-pass2-table}
#| eval: !expr nrow(cna_pass2_long) > 0
#| column: screen-inset

# Display columns for the DT — the 8 raw name_ST19_* / score_ST19_*
# columns are consolidated into a single `ST19_hit` column above (top 3
# score-ordered hits with `parse_st17_pair()`). Raw columns remain in the
# cohort_cna_pass2.tsv output for programmatic consumers.
cna_display_cols <- c(
  "SAMPLE", "sample_id", "driver_tier", "breakpoint", "CytoBand_display", "ID",
  "CNA_TYPE", "WAKHAN_TYPE", "WAKHAN_ID",
  "ALT", "SVTYPE", "SVLEN", "CALLER",
  "TCN", "CN1", "CN2", "CNQ1", "CNQ2",
  "Candidate_genes",
  "Gene_name_display", "ACMG_class", "AnnotSV_ranking_score",
  "P_gain_source", "P_loss_source", "P_ins_source",
  "ST19_hit",
  "BPS", "HET", "GT"
)
cna_display_cols <- intersect(cna_display_cols, names(cna_pass2_clean))

cna_pass2_display <- cna_pass2_clean |>
  arrange(driver_tier, desc(tier_score), CNA_TYPE, SAMPLE, CHROM, POS) |>
  select(all_of(cna_display_cols))

# Reuse the SVTYPE palette for CNA_TYPE (DUP / DEL share colors with SV side).
cna_type_col_js <- which(names(cna_pass2_display) == "CNA_TYPE") - 1L
cna_tier_col_js <- which(names(cna_pass2_display) == "driver_tier") - 1L

# Reuse the SV driver_tier renderer palette (T1 green, T2 orange, T3 grey)
# so the tier badge looks identical across SV / CNA / SNV tables.
cna_tier_render_js <- DT::JS(sprintf(
  "function(data, type, row) {
     if (type !== 'display' || data == null || data === '') return data;
     var colors = %s;
     var c = colors[String(data)] || '#6c757d';
     return '<span style=\"background-color:' + c + ';color:#ffffff;' +
            'font-weight:bold;padding:2px 10px;border-radius:10px;' +
            'display:inline-block;min-width:30px;text-align:center;\">' +
            'T' + data + '</span>';
   }",
  jsonlite::toJSON(list(`1` = "#2CA02C", `2` = "#D68910", `3` = "#6c757d"),
                    auto_unbox = TRUE)
))

cna_type_render_js <- DT::JS(sprintf(
  "function(data, type, row) {
     if (type !== 'display' || data == null || data === '') return data;
     var colors = %s;
     var c = colors[data] || '#6c757d';
     return '<span style=\"background-color:' + c + ';color:#ffffff;' +
            'font-weight:bold;padding:2px 8px;border-radius:4px;' +
            'display:inline-block;min-width:36px;text-align:center;\">' +
            data + '</span>';
   }",
  jsonlite::toJSON(as.list(svtype_colors), auto_unbox = TRUE)
))

DT::datatable(
  cna_pass2_display,
  escape     = FALSE,
  filter     = "top",
  rownames   = FALSE,
  width      = "100%",
  extensions = c("Buttons", "RowGroup"),
  options    = list(
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    pageLength = 15,
    scrollX    = TRUE,
    autoWidth  = TRUE,
    rowGroup   = list(dataSrc = cna_tier_col_js),
    order      = list(list(cna_tier_col_js, "asc")),
    columnDefs = list(
      list(targets = cna_type_col_js, render = cna_type_render_js),
      list(targets = cna_tier_col_js, render = cna_tier_render_js)
    )
  )
)
```

# SNV

## Row {height="120px"}

```{r vb-snv-t1}
#| content: valuebox
#| title: "Tier-1 SNVs"
list(icon = "bullseye", color = "success",
     value = vb_count("snv_long", function(df) df$driver_tier == 1L))
```

```{r vb-snv-t2}
#| content: valuebox
#| title: "Tier-2 SNVs"
list(icon = "geo-alt", color = "primary",
     value = vb_count("snv_long", function(df) df$driver_tier == 2L))
```

```{r vb-snv-t3}
#| content: valuebox
#| title: "Tier-3 SNVs"
list(icon = "paperclip", color = "secondary",
     value = vb_count("snv_long", function(df) df$driver_tier == 3L))
```

```{r vb-snv-candidate}
#| content: valuebox
#| title: "Candidate-gene hits"
list(icon = "diagram-3", color = "info",
     value = vb_count("snv_long", function(df) nzchar(df$Candidate_genes)))
```

## Row {.fill}

### Column {width="65%"}

#### SNV pass2 table {title="SNV pass2 variant table" fill="true"}

```{r snv-status}
if (nrow(snv_long) == 0) {
  cat("**No pass-filtered SNVs found.** Either FILTER_SNV produced no records ",
      "matching the gate for any sample, or the inputs were not staged. ",
      "Table below is suppressed.", sep = "")
}
```

```{r snv-table}
#| eval: !expr nrow(snv_long) > 0
#| column: screen-inset

snv_display_cols <- c(
  "SAMPLE", "sample_id", "driver_tier", "Candidate_genes",
  "CHROM", "POS", "ID", "REF", "ALT", "CALLER",
  "CLAIRSTO_VAF", "DEEPSOMATIC_VAF",
  "SYMBOL", "Consequence", "IMPACT",
  "HGVSc", "HGVSp", "SIFT", "PolyPhen",
  "gnomADg_AF", "CLNSIG", "CLNREVSTAT",
  "SCI", "ONC",
  # CIViC-derived interpretive columns (parsed from CIVIC INFO field).
  # SYMBOL / Consequence appear twice — VEP-derived above, CIViC-derived
  # here — so reviewers can spot annotation disagreements at a glance.
  "CIViC_Consequence", "CIViC_SYMBOL",
  "CIViC_Variant_ID", "CIViC_Variant_URL",
  "CIViC_Entity_Variant_Origin", "CIViC_Entity_Therapies",
  "CIViC_Assertion_ACMG_Codes"
)
snv_display_cols <- intersect(snv_display_cols, names(snv_long))

# DT shows ALL SNVs sorted by driver_tier so T3 records 
snv_display <- snv_long |>
  arrange(driver_tier, desc(tier_score), SAMPLE, CHROM, POS) |>
  select(all_of(snv_display_cols))

snv_tier_col_js <- which(names(snv_display) == "driver_tier") - 1L

# Same green / orange / grey tier badge as SV + CNA tables
snv_tier_render_js <- DT::JS(sprintf(
  "function(data, type, row) {
     if (type !== 'display' || data == null || data === '') return data;
     var colors = %s;
     var c = colors[String(data)] || '#6c757d';
     return '<span style=\"background-color:' + c + ';color:#ffffff;' +
            'font-weight:bold;padding:2px 10px;border-radius:10px;' +
            'display:inline-block;min-width:30px;text-align:center;\">' +
            'T' + data + '</span>';
   }",
  jsonlite::toJSON(list(`1` = "#2CA02C", `2` = "#D68910", `3` = "#6c757d"),
                    auto_unbox = TRUE)
))

DT::datatable(
  snv_display,
  escape = FALSE, filter = "top", rownames = FALSE, width = "100%",
  extensions = c("Buttons", "RowGroup"),
  options = list(
    dom = "Bfrtip", buttons = c("csv", "excel"),
    pageLength = 15, scrollX = TRUE, autoWidth = TRUE,
    rowGroup = list(dataSrc = snv_tier_col_js),
    order = list(list(snv_tier_col_js, "asc")),
    columnDefs = list(list(targets = snv_tier_col_js, render = snv_tier_render_js))
  )
) |>
  # Round every VAF / allele-frequency column to 3 significant digits for
  # readability. Raw values (7+ decimals for gnomAD) survive in the cohort_snv.tsv
  # output for programmatic joins. `intersect(...)` guards against any of
  # these columns being absent in the display frame (e.g. gnomAD not run).
  DT::formatRound(
    columns = intersect(
      c("CLAIRSTO_VAF", "DEEPSOMATIC_VAF", "gnomADg_AF", "gnomADe_AF"),
      names(snv_display)
    ),
    digits = 3
  )
```

### Column {width="35%"}

#### Per-sample SNV summary {height="35%" title="Per-sample SNV summary"}

```{r snv-stats}
#| eval: !expr nrow(snv_long) > 0

snv_stats <- snv_long |>
  group_by(SAMPLE) |>
  summarise(
    `Total SNVs (pass2)`     = n(),
    `HIGH IMPACT`            = sum(IMPACT == "HIGH",     na.rm = TRUE),
    `MODERATE IMPACT`        = sum(IMPACT == "MODERATE", na.rm = TRUE),
    `ClinVar pathogenic`     = sum(grepl("[Pp]athogenic", CLNSIG), na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  snv_stats,
  filter = "top", rownames = FALSE,
  extensions = "Buttons",
  options = list(dom = "Bfrtip", buttons = c("csv", "excel"), pageLength = 25)
)
```

#### Tier-1 lollipop plots {height="65%" title="Tier-1 lollipop plots"}

```{r lollipop-link-table}
#| eval: !expr exists("lollipop_link_display") && is.data.frame(lollipop_link_display) && nrow(lollipop_link_display) > 0

DT::datatable(
  lollipop_link_display,
  escape   = FALSE,
  rownames = FALSE,
  options  = list(dom = "t", pageLength = 30, ordering = FALSE, autoWidth = TRUE)
)
```

```{r lollipop-empty-state, results = "asis"}
#| eval: !expr (!exists("lollipop_link_display")) || (!is.data.frame(lollipop_link_display)) || nrow(lollipop_link_display) == 0

cat("No Tier-1 SNVs with a resolvable gene SYMBOL for lollipop plots.\n")
```







# Information

## Row {height="40%"}

### SV tier scheme {title="SV tier scheme"}

| Tier | Criterion | `tier_score` tie-breaker |
|---|---|---|
| **T1** | Any SVTYPE-matched `name_ST17_*` hit (empirical T-ALL recurrence BED). | max integer score across the SVTYPE-matched `score_ST17_*` cells. |
| **T2** | No ST17 hit, but candidate-gene overlap. | count of overlapping candidate genes. |
| **T3** | Neither ST17 nor candidate-gene, but survived pass2 vembrane gate. | `AnnotSV_ranking_score`. |

### CNA tier scheme {title="CNA tier scheme"}

| Tier | Criterion | `tier_score` tie-breaker |
|---|---|---|
| **T1** | Any of four SVTYPE-matched `name_ST19_Alterations.CNV.Recurrent_*` hits. | max integer score across the four SVTYPE-matched `score_ST19_*` |
| **T2** | Candidate-gene overlap. | count of overlapping candidate genes. |
| **T3** | Passed FILTER_CNA on other evidence. | `AnnotSV_ranking_score`. |

### SNV tier scheme {title="SNV tier scheme"}

| Tier | Criterion | `tier_score` tie-breaker |
|---|---|---|
| **T1** | OncoKB `ONC` = "Oncogenic" AND `SYMBOL` in candidate panel. | max(`CLAIRSTO_VAF`, `DEEPSOMATIC_VAF`). |
| **T2** | ClinVar `CLNSIG` pathogenic-leaning OR `SYMBOL` in candidate panel. | same VAF max. |
| **T3** | Passed FILTER_SNV on other evidence (SIFT/PolyPhen). | same VAF max. |

## Row {height="55%" .tabset}

### SV columns {title="SV column definitions"}

```{r pass2-sv-datadict-btn, results = "asis"}
#| eval: !expr exists("pass2_clean") && nrow(pass2_clean) > 0 && exists("dict_download_button")
cat(dict_download_button("data_dict_sv", "SV"))
```


```{r pass2-sv-datadict, message = FALSE, warning = FALSE}
#| eval: !expr exists("pass2_clean") && nrow(pass2_clean) > 0 && exists("render_data_dict")

# Column-definition table for the SV pass2 DT below. Limited to the columns
# actually shown in the DT (not the full VCF header schema); the "Download
# full data dictionary" button above links to the complete TSV for anyone
# who needs the whole header set.
# display_cols vector must stay in sync with the `pass2-table` chunk.
sv_display_cols <- intersect(c(
  "SAMPLE", "sample_id", "driver_tier", "SVTYPE", "DETAILED_TYPE", "ALT",
  "breakpoint", "CytoBand_display", "has_mate", "SVLEN", "CALLER",
  "Gene_name_display", "Candidate_genes", "ST17_hit",
  "AnnotSV_ranking_score", "ACMG_class",
  "VAF", "hVAF", "DR", "DV", "TUMOUR_READ_SUPPORT", "TUMOUR_AF",
  "ClinVar_SV_CLNSIG"
), names(pass2_clean))

render_data_dict(
  pass2_clean[, sv_display_cols, drop = FALSE], dict_sv,
  caption = "SV column definitions (limited to columns shown in the DT below)"
)
```


### CNA columns {title="CNA column definitions"}

```{r pass2-cna-datadict-btn, results = "asis"}
#| eval: !expr exists("cna_pass2_clean") && nrow(cna_pass2_clean) > 0 && exists("dict_download_button")
cat(dict_download_button("data_dict_cna", "CNA"))
```


```{r pass2-cna-datadict, message = FALSE, warning = FALSE}
#| eval: !expr exists("cna_pass2_clean") && nrow(cna_pass2_clean) > 0 && exists("render_data_dict")

# Column-definition table for the CNA pass2 DT below (columns shown in the DT
# only; the download button above links to the complete VCF header set).
# Kept in sync with the `cna-pass2-table` chunk's cna_display_cols vector.
cna_dict_cols <- intersect(c(
  "SAMPLE", "sample_id", "driver_tier", "breakpoint", "CytoBand_display", "ID",
  "CNA_TYPE", "WAKHAN_TYPE", "WAKHAN_ID",
  "ALT", "SVTYPE", "SVLEN", "CALLER",
  "TCN", "CN1", "CN2", "CNQ1", "CNQ2",
  "Candidate_genes",
  "Gene_name_display", "ACMG_class", "AnnotSV_ranking_score",
  "P_gain_source", "P_loss_source", "P_ins_source",
  "ST19_hit",
  "BPS", "HET", "GT"
), names(cna_pass2_clean))

render_data_dict(
  cna_pass2_clean[, cna_dict_cols, drop = FALSE], dict_cna,
  caption = "CNA column definitions — VCF INFO/FORMAT/FILTER header + qmd-computed columns"
)
```


### SNV columns {title="SNV column definitions"}

```{r snv-datadict-btn, results = "asis"}
#| eval: !expr exists("snv_long") && nrow(snv_long) > 0 && exists("dict_download_button")
cat(dict_download_button("data_dict_snv", "SNV"))
```


```{r snv-datadict, message = FALSE, warning = FALSE}
#| eval: !expr exists("snv_long") && nrow(snv_long) > 0 && exists("render_data_dict")

# Column-definition table for the SNV DT below (columns shown in the DT only;
# the download button above links to the complete VCF header set).
# Kept in sync with the `snv-table` chunk's snv_display_cols vector.
snv_dict_cols <- intersect(c(
  "SAMPLE", "sample_id", "driver_tier", "Candidate_genes",
  "CHROM", "POS", "ID", "REF", "ALT", "CALLER",
  "CLAIRSTO_VAF", "DEEPSOMATIC_VAF",
  "SYMBOL", "Consequence", "IMPACT",
  "HGVSc", "HGVSp", "SIFT", "PolyPhen",
  "gnomADg_AF", "CLNSIG", "CLNREVSTAT",
  "SCI", "ONC",
  "CIViC_Consequence", "CIViC_SYMBOL",
  "CIViC_Variant_ID", "CIViC_Variant_URL",
  "CIViC_Entity_Variant_Origin", "CIViC_Entity_Therapies",
  "CIViC_Assertion_ACMG_Codes"
), names(snv_long))

render_data_dict(
  snv_long[, snv_dict_cols, drop = FALSE], dict_snv,
  caption = "SNV column definitions — VCF INFO/FORMAT/FILTER header + qmd-computed columns"
)
```


## Row {.collapsed}

### Pipeline notes {title="DISCO promotion + citations"}

**Only `driver_tier == 1` records** from all three variant tables are serialized into the DISCO plot's `mlst` array. The cohort DISCO plot on the Overview page shows the union of Tier-1 events across every sample. Per-sample DISCO HTMLs are written to `protein_paint_out/` under `results/sv_report/`.

The `params$candidate_genes` panel and customized BED files from `nextflow.config` are found in `results/annotsv/setupuseranno/` for down-stream investigations.

```{r disco-plot-status}
#| eval: !expr isTRUE(params$disco_enable)

if (!exists("mlst_all") || nrow(mlst_all) == 0) {
  cat("**No Tier-1 variants to display in DISCO plot.** ",
      "The plot is populated from driver_tier == 1 records across SV BNDs, ",
      "CNAs, and SNVs — none survived Tier-1 promotion in this run.", sep = "")
} else {
  n_snv <- sum(mlst_all$dt == 1, na.rm = TRUE)
  n_cna <- sum(mlst_all$dt == 4, na.rm = TRUE)
  n_bnd <- sum(mlst_all$dt == 2, na.rm = TRUE)
  cat(sprintf("**Tier-1 variants displayed:** %d SNV point mutation(s), %d CNA segment(s), %d BND fusion(s). ",
              n_snv, n_cna, n_bnd),
      sprintf("Backend: `%s`. CNA `value` = log2(pmax(TCN, 0.25) / 2) — 0 = diploid, +1 = 4-copy amp, -1 = 1-copy loss.",
              params$proteinpaint_host), sep = "")
}
```


**Citations**

1. Pölönen, P., Di Giacomo, D., Seffernick, A.E. et al. The genomic basis of childhood T-lineage acute lymphoblastic leukaemia. Nature 632, 1082–1091 (2024).
2. Suehnholz, S.P. et al. Quantifying the Expanding Landscape of Clinical Actionability for Patients with Cancer. Cancer Discov 14(1), 49–65 (2024).
3. Ahmad, T. et al. Wakhan: reconstruction of chromosome-scale copy number profiles of tumor genomes with long-read sequencing. medRxiv (2025).
4. Keskus, A.G. et al. Severus detects somatic structural variation and complex rearrangements in cancer genomes using long-read sequencing. Nat Biotechnol 44, 247–257 (2026).
5. Elrick, H. et al. SAVANA: reliable analysis of somatic structural variants and copy number aberrations using long-read sequencing. Nat Methods 22, 1436–1446 (2025).
6. Zheng, Z. et al. ClairS: a deep-learning method for long-read somatic small variant calling. bioRxiv (2023).
7. Park, J. et al. Accurate somatic small variant discovery for multiple sequencing technologies with DeepSomatic. Nat Biotechnol (2025).
